---
title: "Nigeria greta model"
author: ' Ousmane Diao, Nick Golding & Punam Amratia'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    always_allow_html: yes
editor_options: 
  markdown: 
    wrap: 72
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = TRUE,
  comment = "#>",
  message = FALSE
)
```

# epiwave.mapping

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Prototype R package and example code for computationally efficient, fully Bayesian semimechanistic spatio-temporal mapping of disease infection incidence simultaneously from spatially-aggregated (e.g. at health facility level) clinical case count timeseries, and targeted infection prevalence surveys at specific point locations. This model and software is based on ideas in the epiwave R package and uses some functions implemented there, but it currently relies on the user to develop their own model using the greta R package, rather than providing wrapper functions to set up the model for the user.

## Installation

You can install the package from github, using the `remotes` package:

```{r install, eval = FALSE}
remotes::install_github("idem-lab/epiwave.mapping",
                        dependencies = TRUE)
```

## Model

Our main aim is to model, and generate spatio-temporal predictive maps of, variation in *infection incidence*: the average number of new infections per time period, per member of the population. We model the logarithm of infection incidence with a space-time Gaussian process. Since infection incidence is rarely perfectly observed, we infer it from two independent and complementary datastreams:

- **Clinical case timeseries**: Counts of the numbers of cases reported over time, but potentially spatially aggregated across e.g. the catchment areas of health facilities, or the entire region, rather than at spatially precise coordinates

- **Infection prevalence survey points**: Random samples of individuals at a specific point location, with the number tested and number positive recorded. Typically these data are available only very infrequently.

Since neither datastream provides a direct estimate of infection incidence, we model the generative observation process yielding each datastream, using informative priors for key parameters wherever possible. In order to model the full spatial process generating these data, the model requires estimation of the infection incidence across every pixel in a raster grid covering the region of interest - incidence across all cells is aggregated-up to compute the catchment-level case counts. In this way, the model fully accounts for spatial variation in environmental conditions (represented by regression against environmental predictors) and in the residual space-time process across all cells. This model structure leads to a non-linear and non-factorising likelihood which negates many of the computational tricks used in the INLA software for fitting geostatistical models that can be structured as Gaussian Markov random fields. Instead we employ fully Bayesian inference using Hamiltonian Monte Carlo (providing significant flexibility in model specification) and computational approximations to evaluating the Gaussian process across the entire spatial extent, of which a number of options are available.

Below illustrate two methods suited to mapping a separable space-time Gaussian process: simulation with a block-circulant embedding, and with a sparse Gaussian process approximation. The former is similar to the approach used in the `lgcp` R package, and detailed in Diggle et al. (2013). We note that the clinical incidence observation model we employ is a particular case of a Log-Gaussian Cox process. Our contribution is accounting for delays in reporting, and linking model this to prevalence survey data, in a practical and easily-extensible framework.

### Clinical incidence model 

We model the observed number of clinical cases $C_{h,t}$ of our disease of interest in health facility $h$ during discrete time period $t$ as a Poisson random variable:

$$
C_{h,t} \sim \text{Poisson}(\hat{C}_{h,t})
$$
The expectation of this Poisson random variable (the modelled/expected number of clinical cases) is given by a weighted sum of (unobserved but modelled) expected pixel-level clinical case counts $\hat{C}_{l,t}$ at each of $L$ pixel locations $l$:

$$
\hat{C}_{h,t} = \sum_{l=1}^{L}{\hat{C}_{l,t}w_{l,h}}
$$

where weights $w_{l,h}$ give the 'membership' of the population in each pixel location to each health facility, where each case at a given location $l$ has probability $w_{l,h}$ of reporting at health facility $h$, and therefore $\sum_{l=1}^L w_{l,h} = 1$. In practice this could be either a proportional (fractions of the population attend different health facilities) or a discrete (the population of each pixel location attends only one nearby health facility) mapping.

At each pixel location, we model the unobserved clinical case count $\hat{C}_{l,t}$ from the modelled number of new infections $\hat{I}_{l,t'}$ at the same location, during previous time periods $t'$, from the fraction of infections $\gamma_{l,t'}$ at that location and previous time period that would result in a recorded clinical case, and a probability distribution $\pi(t-t')$ over delays $t-t'$ from infection to diagnosis and reporting:

$$
\hat{C}_{l,t} = \sum_{t-t' = 0}^{\tau_\pi}{\hat{I}_{l,t'} \, \gamma_{l,t'} \, \pi(t-t')}
$$
The distribution $\pi(\Delta_t)$ gives the discrete and finite (support on $\Delta_t \in (0, \tau_\pi)$) probability distribution over delays from infection to reporting, indexed on the time-periods considered in the model. This temporal reweighting to account for a distribution over possible delays can be considered as a 'discrete-time convolution' with $\pi(\Delta_t)$ the 'kernel'. Below we discuss efficient methods for computing these convolutions in greta.

### Infection prevalence model 

We model the observed number of individuals who test positive for infection $N^+_{l,t}$ in an infection prevalence survey at location $l$ at time $t$ as a binomial sample, given the number of individuals tested $N_{l,t}$, and the modelled prevalence of infections $\hat{p}_{l,t}$ across the population at that time/place:

$$
N^+_{l,t} \sim \text{Binomial}(N_{l,t}, \hat{p}_{l,t})
$$
Similarly to clinical incidence, we model the infection prevalence at a given location and time as a discrete-time convolution over previous infection counts, divided by the total population in the pixel $M_l$, and the duration of the time-period in days $d$:

$$
\hat{p}_{l,t} = \frac{1}{M_l}\sum_{t-t' = 0}^{\tau_q}{q(t-t') \, \hat{I}_{l,t'} \, d^{-1}}
$$

In this case, the kernel $q(t-t')$ is not a probability distribution, but the kernel of a convolution operation that maps the daily average number of new infections in previous time periods $t'$ to the number of individuals who would test positive (to the diagnostic used) on a randomly-selected day in the timeperiod `t`. This kernel, which is specific to the tim eperiod chosen for modelling the temporal process, can be computed from a related function: $q_{\text{daily}}(\Delta_t)$ which gives the probability that an individual will test positive $\Delta_t$ days after infection. The integral of this function is the average number of days an infected person would test positive for. The function $q_{\text{daily}}()$ can be estimated from empirical data on how the test sensitivity and duration of infection vary over time since infection. This package provides functionality: `transform_convolution_kernel()` to construct timeperiod-specific kernels such as $q()$ from daily kernels like $q_{\text{daily}}()$ for an arbitrary modelling timeperiod.

By convolving of the number of new infections *per day* $\hat{I}_{l,t'} \, d^{-1}$ in all previous time periods with the fraction of those that would test positive in a survey in time period $t$, gives an estimate of the *number* of positive-testing people in the population at location $l$, on any given day within the time period $t$. Dividing this by the population of that location, $M_l$, therefore yields an estimate of the population proportion testing positive on any given day; the parameter of the binomial distribution that captures the prevalence survey data.

Note that our definition of $\hat{p}_{l,t}$ is the population prevalence of infections *that would test positive using that diagnostic method* rather than the true fraction infected/infectious at any one time. We also assume here that tests have perfect specificity, though the model can easily be adapted to situations where that is not the case.

### Infection incidence model 

The expected number of new infections $\hat{I}_{l,t'}$ in location $l$ during time period $t$ is modelled as the product of the population $M_l$ at that location, the length of the timeperiod in days $d$, and the daily infection incidence $f_{l,t}$ at that location and time:

$$
\hat{I}_{l,t'} = d \, M_l \, \hat{f}_{l,t}
$$

<div style="border-left: 5px solid #2c7fb8; padding: 1em; background-color: #f0f8ff;">

<strong>Model Assumption Highlight:</strong>  
Whilst we mechanistically model the observation processes yielding our data types, we employ a <em>geostatistical approach</em> to modelling spatio-temporal variation in infection incidence, with spatio-temporal covariates $\mathbf{X}_{l,t}$, ITN use rates $\mathbf{ITN}_{l,t}$, and a space-time random effect $\epsilon_{l,t} \sim \mathcal{GP}(0, \Sigma)$ with a zero-mean Gaussian process prior.

</div>


$$
\text{log}(f_{l,t}) = \alpha +\mathbf{X}_{l,t} \beta + \epsilon_{l,t}
$$

$$
\epsilon \sim GP(0, \mathbf{K})
$$

where $\alpha$ is a scalar intercept term, $\beta$ is a vector of regression coefficients against the environmental covariates, and $\mathbf{K}$ is the space-time covariance function of the Gaussian process over random effects $\epsilon_{l,t}$.

There are many choices of space-time covariance structure for $\mathbf{K}$, though we use a separable combination of an isotropic spatial covariance function with a temporal covariance function, to enable the use of a range of computationally efficient simulation and calculation methods:

$$
K_{l,t,l',t'} = \sigma^2 \, K_{\text{space}}(||l-l'||; \phi) \, K_{\text{time}}(|t-t'|; \theta)
$$
where $\sigma^2$ is the marginal variance (amplitude) of the resultant Gaussian process, $K_{\text{space}}(.; \phi)$ is an (unit variance) isotropic spatial kernel on euclidean distances $||l-l'||$ with parameter $\phi > 0$ controlling the range of spatial correlation, and $K_{\text{time}}(|t-t'|; \theta)$ is a a temporal kernel on time differences $|t-t'|$, with parameter $\theta$ controlling the range of temporal correlation. Again for computational reasons, we prefer Markovian kernels for $K_{\text{time}}$.

### Infection prevalence - clinical incidence relationship

As described above, the spatially- and temporally-varying parameter $\gamma_{l,t}$  gives the fraction of new infections that would result in a recorded clinical case. In the absence of additional data to that described above, it is unlikely that spatio-temporal variation in this parameter can be inferred. For some diseases - e.g. those where previous infections confer little immune protection from symptoms of future infections - it may be sufficient to model this as a constant, in which case it can be inferred statistically. However for many diseases, there is likely to be a non-linear relationship between the infection incidence in a given population, and the fraction of those infections that result in a clinical case. Where repeat infections decrease the clinical severity of subsequent infections, $\gamma_{l,t}$ is likely to manifest as a monotone decreasing function of infection incidence $f_{l,t}$. In this scenario, it would be preferable to account for this relationship using external information. In the case of malaria, this relationship is often explored using empirical data in terms of the relationship between infection prevalence and clinical incidence. Where a previously-estimated relationship between infection prevalence and clinical incidence is available, the relationship between infection incidence and clinical incidence can be inferred as follows.

We assume that a pre-determined function $g()$ is available that maps from the 'true' population infection prevalence $p_{i}$ to the 'true' underlying clinical incidence $c_{i}$ in some population and setting $i$, for which the corresponding infection incidence is held constant:

$$
c_{i} = g(p_{i})
$$

<div style="border-left: 6px solid #e67e22; background-color: #fff8f0; padding: 1em;">

<p><strong>Reporting Under Imperfect Observation:</strong></p>

<p>In this case, $c_{i}$ represents the true incidence of all clinical episodes, <em>regardless of whether those cases are captured by the health system</em>. We assume that the function $g()$ is estimated using high-quality data sources — such as empirical estimates of $p_{i}$ and $c_{i}$ from intensive surveillance (e.g., prospective cohort studies).</p>

<p>This structure explicitly allows for <strong>imperfect case detection</strong> within routine systems. The observed case counts $C_{i}$, reported through standard health system mechanisms, are assumed to reflect the true burden $c_i$ only partially — mediated by a constant reporting rate $r$:</p>

$$
C_{i} = r \, d \, M_i \, c_{i}
$$

<p>Here, $r$ vary by location to account for differences in health system access or diagnostic capacity to be $r_i$, and $d$ captures the detection probability conditional on treatment-seeking and testing.</p>

</div>

where $d$ is the duration of time periods over which the counts are made, and $M_i$ is the size of this population (converting from clinical incidence to the number of clinical cases, as above). Under the assumption (from $g()$) that infection incidence $f_{i}$ is constant over the period to which $c_{i}$ and $p_{i}$ correspond, these two quantities can therefore be related as:

$$
c_{i} = \kappa_i f_{i}
$$
where $\kappa_i$ is the fraction of infections at location $i$ resulting in a clinical case (recorded or otherwise), and:
$$
p_{i} = q^* \, f_{i}
$$
where $q^*$ is the average number of subsequent days on which each new infection would test positive, if tested once per day (following from the convolution above, in the case that $f_{l,t}$ and therefore $I_{l,t}$ are constant over time); the integral of the detectability function $q()$: 
$$
q^* = \int_{0}^{\tau_q}{q(t) dt}
$$
This scalar value $q^*$ can equally be interpretted as the average number of previous days worth of infections that are detected on the day of a prevalence survey. In practice, $q^*$ can be estimated by a sufficiently-fine resolution discrete sum over $q(t)$.

Combining the above, we have that:

$$
c_i = \kappa_i \, f_i = g(q^* f_i)
$$
and therefore:
$$
\kappa_i = g(q^* f_i) / f_i
$$
simce $\gamma_i = r \, \kappa_i$, and applying this relationship to our spatio-temporally varying infection incidences $f_{l,t}$, we can compute $\gamma_{l,t}$ as:
$$
\gamma_{l,t} = r \, g(q^* f_{l,t}) \, / \, f_{l,t}
$$

where $q^*$ is a scalar constant that can be computed *a priori* from $q$, $g()$ is a deterministic function, also known *a priori*, $f_{l,t}$ is our inference target, the daily infection incidence, and $r$ is the only remaining free parameter in this relationship, which is identifiable in the model.

In practice, the functional form of $g()$ may well permit the computational complexity of this term to be reduced, either by computing a simpler analytic solution, or approximating it with a functionally simpler form with near-equivalent shape. This will likely aid in reducing the computational cost of fitting the model.

## Computational approaches

### Inference

We fit the model via fully Bayesian inference, using MCMC (Hamiltonian Monte Carlo) in the greta R package. Whilst a number of alternative inference algorithms have been proposed and widely adopted for Bayesian and non-Bayesian estimation of spatio-temporal Gaussian process models (e.g. INLA), in out case, the aggregation of clinical cases across catchments (a feature required by the model structure) leads to a non-factorising likelihood, which nullifies many of the computational advantages of that method. Further, the mapping from the Gaussian process to the expectations of the two sampling distributions is non-linear (due to sums of exponents in various places), requiring computationally costly linearisation of the non-linearities. By employing a fully-Bayesian MCMC approach, and focussing on computationally-efficient model and implementation choices, rather than approximations, we are able to estimate the model parameters relatively quickly, with full treatment, propagation, and quantification of model uncertainty, and retaining the ability to easily modify and interrogate the model.

### Discrete-time convolution

This discrete convolutions (weighted sums over previous time points) used to compute $\hat{C}_{l,t}$ and $\hat{p}_{l,t}$ can be computationally intensive, depending on the number of time periods modelled. A number of efficient computational approaches exist to overcome these, and the optimal approach to use in greta depends on the size of $\tau^{max}$: the maximum duration of the delay in terms of the number of timeperiods being considered. Where $\tau^{max}$ is relatively large (say, $\tau^{max} > 3$), implementation of the convolution as a matrix multiply is likely to be the most efficient in greta. If $\tau^{max} > 3$ and the number of time periods being modelled is much larger than $\tau^{max}$, implementing this as a sparse matrix multiply (skipping computation on zero elements of the convolution matrix) is likely to be most efficient. Where $\tau^{max}$ is small (say, $\tau^{max} \leq 3$) the convolution can instead be computed with a sum of dense vectorised additions and subtractions. These convolution approaches are implemented in this package, and demonstrated below.

### Gaussian process simulation

Naive implementation of the full Gaussian process (GP) model is typically very computationally intensive, due to the fact that the most expensive step (inverting a dense covariance matrix) scales cubically with the number of evaluation points. In the model we propose (and as in a Log-Gaussian Cox process), we need to evaluate the GP at every pixel location and time period in the study frame, so that we can aggregate the expected clinical case count to compute the likelihood. This results in a computationally impractical algorithm for all but the smallest study frames. Some form of computational approximation is required.

One straightforward approximation would be to employ a full (non-approximate) GP, but to evaluate the combined process (GP and fixed effects regression) at only a subset of locations and times in the study frame and then to approximate the health-facility level clinical case count calculation (a set of integrals over space) with some smaller finite sum. Whilst general, this would also result in an approximation to the fixed-effects environmental regression part of the model. While the GP could feasibly be smooth and well approximated in this way, the environmental covariates are likely to be more complex and thus potentially quite poorly approximated by a finite sum.

An alternative approach would be to evaluate the GP and the fixed effects at all pixels, but to replace the full GP with an approximation. Candidate GP approximation approaches include the SPDE approximation to a Matern-type spatial kernel, on a computational mesh (as used in INLA, [Lindgren *et al.*, 2011](https://doi.org/10.1111/j.1467-9868.2011.00777.x)), a sparse GP method over a limited set of inducing points ([Quinonera-Candela & Rasmussen, 2005](https://jmlr.org/papers/v6/quinonero-candela05a.html) - see `greta.gp` and note the similarity to a Guassian predictive process per [Banerjee *et al.*, 2009](https://doi.org/10.1111/j.1467-9868.2008.00663.x)), or one of the closely-related penalised spline methods (see `greta.gam`). Implementation of the SPDE approximation in an HMC setting requires implementation of sparse matrix methods adapted to automatic differentiation ([Durrande *et al.* 2019](https://doi.org/10.48550/arXiv.1902.10078), greta version in development), but would likely be preferable where the underlying Gaussian process has short lengthscale. In the case of a comparatively smooth underlying GP, the sparse o spline-based methods are likely to be a better fit, since they have lower computational complexity in this case. Given the spatially-aggregated observation model, it seems likely that only comparatively smooth GP realisations could be identified in this model, and therefore favour sparse or spline methods. 

An appealing third alternative in this case is to compute the full spatial GP for every pixel in a regular grid (the same we use to record spatial covariate values), by exploiting the block-circulant structure in the resulting spatial covariance matrix. This requires using projected coordinates, and expanding the spatial study frame (by a factor of 2 in each dimension) to map the projection onto a torus in such a way that the distances between pairs of pixels are preserved. This approach enables simulation of the GP across all pixels, for a given time period, via the fast fourier transform (FFT) - which scales only linearly with the number of locations considered (? check when have wifi). This enables us to simulate values of the spatial GP across all pixels very cheaply, with no need to approximate the GP. In separable combination with a Markovian temporal kernel, this enables very rapid inference. However the very large number of random variables required to be sampled (at least four times as many as pixels and time points under study) leads to a very large parameter space for sampling, and can slow down efficient MCMC sampling considerably.

## Example application

We demonstrate the model with application to mapping the (simulated) infection incidence of malaria in Kenya from simulated clinical incidence and infection prevalence data. Using the `sim_data()` function from this package, we use a real grid of environmental covariates, but simulate the locations of health facilities, prevalence survey locations, and all datasets to which we will then fit the model.

Define reference image, population and admin masks ----
# Aggregate to 5km resolution (factor = 5)

```{r, cache=FALSE}
#rm(list = ls())

library(sf)
library(terra)
#library(raster)

pop_u5_orig <- rast("/mnt/s3/stratification/CountryFolders/Nigeria/Common files/Population/u5_grid3_1km_new.tif")
pop_u5_agg <- aggregate(pop_u5_orig, fact = 5, fun = sum, na.rm = TRUE)
# 1. Read shapefile (still in WGS84)
shape <- st_read("/mnt/s3/stratification/CountryFolders/Nigeria/Common files/Shapefiles/NGA_LGAs.shp") %>%
  st_make_valid()
# 2. Transform shapefile to match raster CRS
shape <- st_transform(shape, crs(pop_u5_agg))
# 3. Convert sf to terra SpatVector
shape_vect <- vect(shape)
# 4. Mask raster using boundary
pop_masked <- mask(pop_u5_agg, shape_vect)

# 5. Get valid cell indices where population > 0
valid_cells <- which(!is.na(values(pop_masked)) & values(pop_masked) > 0)
valid.pix     <- valid_cells
# 6. Extract population values and coordinates
population <- values(pop_masked)[valid_cells]
raster_coords <- terra::xyFromCell(pop_masked, valid_cells)
valid.coords <- raster_coords
# Change projection to UTM
new_crs <-"epsg:26332"  #"epsg:21036"
pop_u5 <- terra::project(pop_masked, new_crs)
```

The Gaussian process implementations we use below assume that the coordinates are on a plane - they don't account for the curvature of the earth (unlike INLA which is able to model GPs on a sphere). We therefore first need to ensure that the raster we use to define this computational setup is appropriately projected. For the block-circulant matrix approach, this is especially important since the computation relies on the compute locations being arranged in a regular gird. We therefore project our matrix to an appropriate projected coordinate reference system:

```{r }
library(terra)
# define a template raster representing our compute grid: 0 in land cells, NA
# elsewhere
grid_raster <- pop_u5 * 0
names(grid_raster) <- NULL

# get an index to all pixel values
all_cells <- terra::cells(grid_raster)
```

Load static covariates
```{r load_static_covs, cache=TRUE}
#load("covs_static.Rdata")

# Built covariates as raster stack
# Read and process each raster
dir_path <- "/mnt/s3/stratification/CountryFolders/Nigeria"
raster_dir <- paste0(dir_path,"/Common files/Covariates/1km/")
library(terra)

dir_path <- "/mnt/s3/stratification/CountryFolders/Nigeria"
raster_dir <- paste0(dir_path,"/Common files/Covariates/1km/")
covariate.names <- list(
  "accessibility_to_cities_2015_v1.0.tif",
  "Aridity_Index_v2.Synoptic.Overall.Data.1km.Data.tif",
  "distance_to_water.tif",
  "SRTM_Elevation.Synoptic.Overall.Data.1km.mean.tif", #replace by MERIT
  "PET_v2.Synoptic.Overall.Data.1km.Data.tif",
  "SRTM_SlopePCT_Corrected.Synoptic.Overall.Data.1km.Data.tif",
  "DMSP.F18_Stable_Lights_v4c.2013.Annual.Data.1km.Data.tif"
)

Ncovariates <- length(covariate.names)
cov_stack <- list()

# Create 5km template from pop_u5
ref_template <- pop_u5  # Use this as the 5km grid

for (k in 1:Ncovariates) {
  cat("Processing:", covariate.names[[k]], "\n")

  cov_current <- rast(paste0(raster_dir, covariate.names[[k]]))
  
  # Project and align to pop_u5 (1km) first
  cov_current <- terra::project(cov_current, crs(pop_u5))
  cov_current <- crop(cov_current, pop_u5)
  cov_current <- extend(cov_current, pop_u5)
  
  # Aggregate to 5km resolution
  cov_5km <- aggregate(cov_current, fact = 5, fun = mean, na.rm = TRUE)
  
  # Resample to match final 5km grid exactly
  cov_5km <- resample(cov_5km, ref_template)
  
  cov_stack[[k]] <- cov_5km
}

# Combine into a 5km SpatRaster stack
covariates_raster <- rast(cov_stack)
#names(covariates_raster) <- paste0("cov_", seq_len(Ncovariates))

# Scale each layer
covariates_raster <- scale(covariates_raster)
```

Spatial nearest-neighbor imputation in order to avoid NAs in my covariates
```{r extract_static_covs}
# Extract covariate values at modeling locations (e.g., all_cells)
X <- terra::extract(covariates_raster, all_cells)

# Impute NA in each column
for (j in seq_len(ncol(X))) {
  nas <- which(is.na(X[, j]))
  if (length(nas) > 0) {
    valid <- which(!is.na(X[, j]))
    for (i in nas) {
      nearest <- which.min(
        (valid.coords[valid, 1] - valid.coords[i, 1])^2 +
        (valid.coords[valid, 2] - valid.coords[i, 2])^2
      )
      X[i, j] <- X[valid[nearest], j]
    }
  }
}

X[X > 3] <- 3 #truncate to mimic non-linearity in the extreme
X[X < -3] <- -3 #truncate to mimic non-linearity in the extreme
```

Load dynamic covariates
```{r load_dynamic_covs, cache=TRUE}

load_dynamic_covariate <- function(varname, startyear, endyear, ref_raster, new_crs, fun = mean) {
  cov_stack <- list()
  k <- 1
  for (year in startyear:endyear) {
    for (month in 1:12) {
      fpath <- sprintf("~/epiwave.mapping/Dynamic_covariates_1km/%s_v061.%d.%02d.mean.1km.Data.tif", varname, year, month)
      if (!file.exists(fpath)) next
      r <- rast(fpath)
      r <- project(r, new_crs)
      r <- crop(r, pop_u5)
      r <- extend(r, pop_u5)
      r <- aggregate(r, fact = 5, fun = fun, na.rm = TRUE)
      r <- resample(r, ref_raster)
      cov_stack[[k]] <- r
      k <- k + 1
    }
  }
  rast(cov_stack)
}
```

```{r scale_dynamic_covs}
# Load and scale
EVI_covariates_raster       <- scale(load_dynamic_covariate("EVI", 2021, 2023, pop_u5, new_crs))
LST_Day_covariates_raster   <- scale(load_dynamic_covariate("LST_Day", 2021, 2023, pop_u5, new_crs))
LST_Night_covariates_raster <- scale(load_dynamic_covariate("LST_Night", 2021, 2023, pop_u5, new_crs))
TCB_covariates_raster       <- scale(load_dynamic_covariate("TCB", 2021, 2023, pop_u5, new_crs))
TCW_covariates_raster       <- scale(load_dynamic_covariate("TCW", 2021, 2023, pop_u5, new_crs))
```
Extrating the dynamic covariates

```{r}

EVI_temporal_cov <- terra::extract(EVI_covariates_raster, all_cells)
# Impute NA in each column
for (j in seq_len(ncol(EVI_temporal_cov))) {
  nas <- which(is.na(EVI_temporal_cov[, j]))
  if (length(nas) > 0) {
    valid <- which(!is.na(EVI_temporal_cov[, j]))
    for (i in nas) {
      nearest <- which.min(
        (valid.coords[valid, 1] - valid.coords[i, 1])^2 +
        (valid.coords[valid, 2] - valid.coords[i, 2])^2
      )
      EVI_temporal_cov[i, j] <- EVI_temporal_cov[valid[nearest], j]
    }
  }
}
EVI_temporal_cov[EVI_temporal_cov > 3] <- 3 
EVI_temporal_cov[EVI_temporal_cov < -3] <- -3 
```

```{r}
LST_day_temporal_cov <- terra::extract(LST_Day_covariates_raster, all_cells)
# Impute NA in each column
for (j in seq_len(ncol(LST_day_temporal_cov))) {
  nas <- which(is.na(LST_day_temporal_cov[, j]))
  if (length(nas) > 0) {
    valid <- which(!is.na(LST_day_temporal_cov[, j]))
    for (i in nas) {
      nearest <- which.min(
        (valid.coords[valid, 1] - valid.coords[i, 1])^2 +
        (valid.coords[valid, 2] - valid.coords[i, 2])^2
      )
      LST_day_temporal_cov[i, j] <- LST_day_temporal_cov[valid[nearest], j]
    }
  }
}
LST_day_temporal_cov[LST_day_temporal_cov>3] <- 3
LST_day_temporal_cov[LST_day_temporal_cov<-3] <- -3
```

```{r}
LST_night_temporal_cov <- terra::extract(LST_Night_covariates_raster, all_cells)
# Impute NA in each column
for (j in seq_len(ncol(LST_night_temporal_cov))) {
  nas <- which(is.na(LST_night_temporal_cov[, j]))
  if (length(nas) > 0) {
    valid <- which(!is.na(LST_night_temporal_cov[, j]))
    for (i in nas) {
      nearest <- which.min(
        (valid.coords[valid, 1] - valid.coords[i, 1])^2 +
        (valid.coords[valid, 2] - valid.coords[i, 2])^2
      )
      LST_night_temporal_cov[i, j] <- LST_night_temporal_cov[valid[nearest], j]
    }
  }
}
LST_night_temporal_cov[LST_night_temporal_cov>3] <- 3
LST_night_temporal_cov[LST_night_temporal_cov<-3] <- -3
```

```{r}
TCB_temporal_cov <- terra::extract(TCB_covariates_raster, all_cells)
# Impute NA in each column
for (j in seq_len(ncol(TCB_temporal_cov))) {
  nas <- which(is.na(TCB_temporal_cov[, j]))
  if (length(nas) > 0) {
    valid <- which(!is.na(TCB_temporal_cov[, j]))
    for (i in nas) {
      nearest <- which.min(
        (valid.coords[valid, 1] - valid.coords[i, 1])^2 +
        (valid.coords[valid, 2] - valid.coords[i, 2])^2
      )
      TCB_temporal_cov[i, j] <- TCB_temporal_cov[valid[nearest], j]
    }
  }
}
TCB_temporal_cov[TCB_temporal_cov>3] <- 3
TCB_temporal_cov[TCB_temporal_cov<-3] <- -3
```

```{r}
TCW_temporal_cov <- terra::extract(TCW_covariates_raster, all_cells)
# Impute NA in each column
for (j in seq_len(ncol(TCW_temporal_cov))) {
  nas <- which(is.na(TCW_temporal_cov[, j]))
  if (length(nas) > 0) {
    valid <- which(!is.na(TCW_temporal_cov[, j]))
    for (i in nas) {
      nearest <- which.min(
        (valid.coords[valid, 1] - valid.coords[i, 1])^2 +
        (valid.coords[valid, 2] - valid.coords[i, 2])^2
      )
      TCW_temporal_cov[i, j] <- TCW_temporal_cov[valid[nearest], j]
    }
  }
}
TCW_temporal_cov[TCW_temporal_cov>3] <- 3
TCW_temporal_cov[TCW_temporal_cov<-3] <- -3
```

```{r load_chirps}
#----------------------------CHIRPS
Nyears = 3
startyear = 2021
endyear = 2023
ref_template <- pop_u5  # Use this as the 5km grid
cov_stack <- list()
k <- 1
for (year in startyear:endyear) {
  cat(paste0(year, ": "))
  for (month in 1:12) {
    file_path <- paste0("~/epiwave.mapping/Dynamic_covariates_1km/chirps-v2-0.",year,".",sprintf("%02i",month),".sum.1km.NN.tif")
    
    cov.current <- rast(file_path)
    cov.current <- terra::project(cov.current, new_crs)
    cov.current <- crop(cov.current, pop_u5)
    cov.current <- extend(cov.current, pop_u5)
    
    cov_5km <- aggregate(cov.current, fact = 5, fun = sum, na.rm = TRUE)
    cov_5km <- resample(cov_5km, ref_template)

    cov_stack[[k]] <- cov_5km
    k <- k + 1
    cat(month.abb[month], " ")
  }
  cat("\n")
}

# Combine into a 5km SpatRaster stack
CHIRPS_covariates_raster <- rast(cov_stack)
# Scale each layer
CHIRPS_covariates_raster <- scale(CHIRPS_covariates_raster)
```

```{r}
CHIRPS_temporal_cov <- terra::extract(CHIRPS_covariates_raster, all_cells)
# Impute NA in each column
for (j in seq_len(ncol(CHIRPS_temporal_cov))) {
  nas <- which(is.na(CHIRPS_temporal_cov[, j]))
  if (length(nas) > 0) {
    valid <- which(!is.na(CHIRPS_temporal_cov[, j]))
    for (i in nas) {
      nearest <- which.min(
        (valid.coords[valid, 1] - valid.coords[i, 1])^2 +
        (valid.coords[valid, 2] - valid.coords[i, 2])^2
      )
      CHIRPS_temporal_cov[i, j] <- CHIRPS_temporal_cov[valid[nearest], j]
    }
  }
}
CHIRPS_temporal_cov[CHIRPS_temporal_cov>3] <- 3
CHIRPS_temporal_cov[CHIRPS_temporal_cov<-3] <- -3
```

Load treatment seeking rate
```{r all_s_ts_matrix}
#load("all_s_ts.Rdata")
cov_current <- rast("/mnt/s3/stratification/CountryFolders/Nigeria/Projects/1.0_HBHI_risk_mapping/04 Models/jan2023-update/treatment-seeking/2021_ts_dhis2_1km.tif")

# Step 2: Make sure both rasters have the same CRS (coordinate reference system)
if (crs(cov_current) != crs(pop_u5)) {
  cov_current <- terra::project(cov_current, new_crs)  # Reproject all_s_ts to match buffer.image's CRS
}
# Step 3: Crop or Extend
# If you want to crop all_s_ts to the extent of buffer.image
cov_current <- crop(cov_current, pop_u5)  # Crop to the extent of buffer.image

# If you want to extend all_s_ts to match the extent of buffer.image
cov_current <- extend(cov_current, ext(pop_u5))  # Extend to the extent of buffer.image

cov_5km <- aggregate(cov_current, fact = 5, fun = mean, na.rm = TRUE)
cov_5km <- resample(cov_5km, ref_template)

TS_covariates_raster <- cov_5km
```

```{r}

all_s_ts <- terra::extract(TS_covariates_raster, all_cells)
all_s_ts_vec <- all_s_ts[[1]]  # extract the vector of values
nas <- which(is.na(all_s_ts_vec))
if (length(nas) > 0) {
  valid <- which(!is.na(all_s_ts_vec))
  for (i in nas) {
    nearest <- which.min(
      (valid.coords[valid, 1] - valid.coords[i, 1])^2 +
      (valid.coords[valid, 2] - valid.coords[i, 2])^2
    )
    all_s_ts_vec[i] <- all_s_ts_vec[valid[nearest]]
  }
}
```

Load ITN use rate
```{r ITN_use_matrix}
#load("ITN_use.RData")
library(terra)
#ITN_temporal_cov_5km <- matrix(0, nrow = length(valid.pix), ncol = 12)
cov_stack <- list()
k<-1
for (month_idx in 1:12) {
  fpath <- paste0("/mnt/s3/stratification/CountryFolders/Nigeria/Common files/Covariates/1km/Monthly/ITN/20210531_v1/ITN_use_1km_monthly_", month_idx, ".tif")

  if (!file.exists(fpath)) {
    warning("Missing file: ", fpath)
    next
  }

  cov_current <- rast(fpath)
  cov_current <- terra::project(cov_current, new_crs)
  cov_current <- crop(cov_current, pop_u5)
  cov_current <- extend(cov_current, pop_u5)

  # Aggregate to 5km
  cov_5km <- aggregate(cov_current, fact = 5, fun = mean, na.rm = TRUE)
  cov_5km <- resample(cov_5km, pop_u5)

  cov_stack[[k]] <- cov_5km
  k <- k + 1
  cat(month.abb[month_idx], " ")
}

# Combine into a 5km SpatRaster stack
ITN_covariates_raster <- rast(cov_stack)
```

```{r}
ITN_temporal_cov <- terra::extract(ITN_covariates_raster, all_cells)
# Impute NA in each column
for (j in seq_len(ncol(ITN_temporal_cov))) {
  nas <- which(is.na(ITN_temporal_cov[, j]))
  if (length(nas) > 0) {
    valid <- which(!is.na(ITN_temporal_cov[, j]))
    for (i in nas) {
      nearest <- which.min(
        (valid.coords[valid, 1] - valid.coords[i, 1])^2 +
        (valid.coords[valid, 2] - valid.coords[i, 2])^2
      )
      ITN_temporal_cov[i, j] <- ITN_temporal_cov[valid[nearest], j]
    }
  }
}
ITN_temporal_cov_36 <- do.call(cbind, replicate(3, ITN_temporal_cov, simplify = FALSE))
```

Plot the population
```{r pop_plot}
library(terra)
# Ensure pop_raster is a SpatRaster
#Convert SpatRaster to Data Frame
pop_raster_df <- as.data.frame(pop_u5, xy = TRUE)

library(ggplot2)
ggplot()+
  geom_tile(data=pop_raster_df, aes(x = x, y = y,fill = u5_grid3_1km_new))+
  #use na.value to make the lakes appear as lightblue
  scale_fill_viridis_c(option = "magma", na.value = "lightblue", trans = 'log10')+
  theme_minimal()+
  labs(fill = "Population - U5 in Nigeria", fill = "Population Density")

```


```{r TS-plots}
library(ggplot2)
all_s_ts_orig <- rast("/mnt/s3/stratification/CountryFolders/Nigeria/Projects/1.0_HBHI_risk_mapping/04 Models/jan2023-update/treatment-seeking/2021_ts_dhis2_1km.tif")

shape$alls_ts <- exactextractr::exact_extract(all_s_ts_orig,shape, 'mean')
ggplot(shape)+
  geom_sf(mapping = aes(fill = alls_ts))+
  scale_fill_distiller(palette = "BrBG", direction = 1)+
  labs(title="Treatment seeking DHIS2 only in 2021")+
  theme(strip.text = element_text(size = 16, face = "bold"),
        plot.title = element_text(size = 16, face = "bold"),   # Title size
        plot.caption = element_text(size = 12),               # Caption size
        axis.title.x = element_text(size = 16),  # X-axis label size
        axis.title.y = element_text(size = 16),  # Y-axis label size
        axis.text.x = element_text(size = 14),   # X-axis tick size
        axis.text.y = element_text(size = 14),   # Y-axis tick size
        legend.text = element_text(size = 14),                # Legend text size
        legend.title = element_text(size = 14, face = "bold"), # Legend title size
        legend.position = 'bottom',
        legend.key.width = unit(2., "cm"),                  # Largeur des clés
        legend.key.height = unit(0.5, "cm"))
```



Read and plot ITN use in LGA level
```{r ITN_use_plots}
library(raster)
library(terra)
library(tidyverse)
itn_stack <- stack(list.files("/mnt/s3/stratification/CountryFolders/Nigeria/Common files/Covariates/1km/Monthly/ITN/20210531_v1/", ".tif", full.names = TRUE))
names(itn_stack) <- month.name[as.numeric(sub("ITN_use_1km_monthly_", "", names(itn_stack)))]
# Select only layers with valid month names (January - December)
months <- c("January", "February", "March", "April", "May", "June", 
            "July", "August", "September", "October", "November", "December")

# Extract matching layers
itn_stack <- itn_stack[[which(names(itn_stack) %in% months)]]

new_crs <-"epsg:26332"  #"epsg:21036"
# project these file
itn_stack <- terra::project(rast(itn_stack), new_crs)

itn_monthly <- terra::extract(itn_stack, shape, 'mean')
 names(itn_monthly) = sub("mean.", "", names(itn_monthly))
# # 
itn_monthly <- as_data_frame(itn_monthly) %>% 
  dplyr::select(January,February,March,April, May, June,July,August,    September,October,November,December)

itn_map <- dplyr::bind_cols(shape, itn_monthly)

# Ensure month labels are correct
month_labels <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

# Convert dataset to long format
itn_map_long <- itn_map %>%
  pivot_longer(cols = 11:22, names_to = "month", values_to = "ITN_Use") %>%
  mutate(month = factor(month, levels = names(itn_map)[11:22], labels = month_labels))


ggplot(itn_map_long)+
  geom_sf( aes(geometry=geometry, fill=ITN_Use))+
  scale_fill_distiller(palette = "Spectral", na.value = "lightblue")+
  facet_wrap(~month)+
  labs(fill = "Annual ITN use ", title = "Nigeria LGAs", subtitle = "MIS 2020, 0-5 years")+
  theme_bw()
```


Plot the clinical case timeseries for all health facilities over each year:

```{r clinical_cases_observed}
library(tidyverse)
nga_inc_data <- read.csv("/mnt/efs/stratification/Nigeria/HBHI_2025/03_Additional_files_optional/Imputation/tgo_dhis2_imputed_aggregated.csv")%>%
  arrange(date) %>%
  separate(date, into = c('monthname', 'year'), sep = " ") %>% 
  mutate(month = factor(monthname, levels = c("Jan","Feb","Mar","Apr","May","Jun",
                                              "Jul","Aug","Sep","Oct","Nov","Dec")),
         month = as.numeric(month),
         year = as.numeric(year)) 

cases <- nga_inc_data 
cases <- cases %>% filter(!is.na(positive))
ggplot(
  aes(
    x = month,
    y = positive,
    colour = adm2,
    group = adm2
  ),
  data = cases) +
  facet_wrap(~year) +
  geom_line(
    linewidth = 0.3
  ) + 
  theme_minimal() +
  scale_colour_discrete(
    guide = "none"
  ) +
  scale_x_continuous(breaks = 1:12) +
  ggtitle("Clinical case timeseries")
#ggsave("~/epiwave.mapping/clinical_cases_timeseries.png")
```
```{r agg_cases_observed}
library(sf)
library(dplyr)
library(ggplot2)

nga_lga <- sf::st_read(paste0("/mnt/s3/stratification/CountryFolders/Nigeria/Common files/Shapefiles/NGA_clean_Shp/NGA_LGAs_fix/NGA_LGAs3.shp"), quiet = TRUE) %>%
  mutate(
    adm1 = snakecase::to_snake_case(State),
    adm2 = snakecase::to_snake_case(LGA)
  ) %>%
  dplyr::select(adm1, adm2, geometry)

# Join data to shapefile
map_data <- left_join(nga_lga, nga_inc_data, by = "adm2")

# Plot faceted map by year
ggplot(map_data) +
  geom_sf(aes(fill = positive), color = "white") +
  facet_wrap(~ year) +
  scale_fill_distiller(palette = "RdYlGn", na.value = "lightblue", name = "Cases")+
  #scale_fill_viridis_c(option = "C", na.value = "gray90", name = "Cases") +
  labs(title = "Malaria Positivity by LGA and Year") +
  theme_minimal()

```



Plot the prevalence survey results at the prevalence survey locations:
read prevalence data
```{r}
library(tidyverse)
prev_data <- read_csv("/mnt/s3/stratification/CountryFolders/Nigeria/Projects/1.0_HBHI_risk_mapping/02 Cleaned data/PfPR_NGA_2023.csv") %>% #keep DHS data only...
  janitor::clean_names() %>% 
  mutate(date = make_date(year = year, month = month))

```

```{r prevalence_mic}
prev_surveys_mic <- prev_data %>%
  rowwise()%>%
   mutate(n_positive=malaria_test_micro,
          n_sampled = n,
     prevalence = n_positive / n_sampled
   )
# 
 ggplot() +
   geom_sf(data = shape) +
   scale_fill_gradient(
     low = grey(0.9),
     high = grey(0.9),
     na.value = "transparent",
     guide = "none") +
   geom_point(
     aes(
       y = latitude,
       x = longitude,
       colour = prevalence
     ),
     data = prev_surveys_mic,
     size = 2,
     shape = 16
   ) +
   scale_colour_gradient(
     labels = scales::label_percent(),
     low = "skyblue",
     high = "darkred"
   ) +
   xlab("") +
   ylab("") +
   theme_minimal() +
   ggtitle("Prevalence surveys with PCR test",
           subtitle="MIS-2021 0-5 years in Nigeria")
```

```{r prevalence_rdt}
prev_surveys_rdt <- prev_data %>%
  rowwise()%>%
   mutate(n_positive=malaria_test_rapid,
          n_sampled = n,
     prevalence = n_positive / n_sampled
   )
# 
 ggplot() +
   geom_sf(data = shape) +
   scale_fill_gradient(
     low = grey(0.9),
     high = grey(0.9),
     na.value = "transparent",
     guide = "none") +
   geom_point(
     aes(
       y = latitude,
       x = longitude,
       colour = prevalence
     ),
     data = prev_surveys_rdt,
     size = 2,
     shape = 16
   ) +
   scale_colour_gradient(
     labels = scales::label_percent(),
     low = "skyblue",
     high = "darkred"
   ) +
   xlab("") +
   ylab("") +
   theme_minimal() +
   ggtitle("Prevalence surveys with RDT test",
           subtitle="MIS-2021 for 0-5 years in Nigeria")
#ggsave("~/epiwave.mapping/Prevalence_rdt.png")
```
```{r background_fever}
library(tidyverse)
bg_data <- read_csv("/mnt/s3/stratification/CountryFolders/Nigeria/Projects/1.0_HBHI_risk_mapping/02 Cleaned data/BG_NGA_2023.csv") %>% #keep DHS data only...
  janitor::clean_names() %>%
  mutate(date = make_date(year = year, month = month))
bg_data <- filter(bg_data, latitude != 0, longitude != 0)

bg_data<- bg_data %>% 
  rowwise()%>%
  mutate(
         positive = fever,
         tested=n,
         rate = fever/tested)
library(ggplot2)
 ggplot() +
   geom_sf(data = shape) +
   scale_fill_gradient(
     low = grey(0.9),
     high = grey(0.9),
     na.value = "transparent",
     guide = "none") +
   geom_point(
     aes(
       y = latitude,
       x = longitude,
       colour = rate
     ),
     data = bg_data,
     size = 2,
     shape = 16
   ) +
   scale_colour_gradient(
     labels = scales::label_percent(),
     low = "skyblue",
     high = "darkred"
   ) +
   xlab("") +
   ylab("") +
   theme_minimal() +
   ggtitle("Background of fever surveys",
           subtitle="MIS-2021 for 0-5 years in Nigeria")
#ggsave("~/epiwave.mapping/Background-fever.png")
```

```{r}
n_months<- dim(EVI_temporal_cov)[2]
save(
  all_cells,
  covariates_raster,
  EVI_temporal_cov,
  LST_day_temporal_cov,
  LST_night_temporal_cov,
  CHIRPS_temporal_cov,
  TCB_temporal_cov,
  TCW_temporal_cov,
  ITN_temporal_cov_36,
  all_s_ts_vec,
  pop_u5,
  grid_raster,
  population,
  valid.coords,
  nga_inc_data,
  n_months,
  file = "model.inputs.Rdata"
)
```
Plot the fake delay distribution (probability mass function over days) and prevalence survey detectability functions:

```{r Case_reporting_delay_distribution}
max_case_delay <- 14
pi_daily <- function(day_difference) {
    upper_limit <- max_case_delay
    # use a truncated lognormal (truncated to 0-upper_limit)
    mu <- 1.2
    sigma <- 0.5
    day_difference_old <- day_difference
    day_difference <- round(day_difference)
    if (!max(abs(day_difference - day_difference_old) < 1e-4)) {
      warning("day_difference was non-integer, using rounded values")
    }

    upper <- plnorm(day_difference + 1, mu, sigma)
    lower <- plnorm(day_difference, mu, sigma)
    dens <- upper - lower

    # correct the densities
    norm <- plnorm(upper_limit, mu, sigma)
    res <- dens / norm
    res[day_difference > upper_limit] <- 0
    res[day_difference < 0] <- 0
    res
  }

case_delay_plot <- tibble(
    days = seq(0, max_case_delay)) %>%
    mutate(pmf = pi_daily(days)
   )
# 
 ggplot(
   aes(x = days,
       y = pmf),
   data = case_delay_plot) +
   geom_step() +
   theme_minimal() +
   ggtitle("Case reporting delay distribution",
           "Probability of reporting X days after infection")
```

```{r Prevalence_survey_detectability_function}
prev_detectability_max_days <- 30
q_daily <- function(days) {
    up <- plogis((days * 2) - 5)
    down <- 1 - plogis(days / 2 - 10)
    q <- up  * down
    q[days < 0] <- 0
    q[days > 30] <- 0
    q
  }
 detectability_plot <- tibble(
   days = seq(0, prev_detectability_max_days)) %>%
   mutate(
     detectability = q_daily(days)
   )
# 
 ggplot(
   aes(x = days,
       y = detectability),
   data = detectability_plot) +
   geom_step() +
   theme_minimal() +
   ggtitle("Prevalence survey detectability function",
           "Probability of registering a positive if tested X days after infection")
```

Plot the assumed infection prevalence - clinical case incidence relationship:

```{r Prevalence-clinical_incidence_relationship}
q_max_days <- 30
q_star <- sum(q_daily(seq(0, q_max_days)))

g <- function(prevalence) {
    (1/365) * (1 - exp(-prevalence * 10))
  }

prev_inc_plot <- tibble(prevalence = prev_surveys_rdt$prevalence) %>%
  filter(prevalence >= 0 & prevalence <= 1 & !is.na(prevalence)) %>%
  mutate(clinical_incidence = g(prevalence))
# 
 ggplot(
   aes(x = prevalence,
       y = 365 * clinical_incidence),
   data = prev_inc_plot) +
   geom_line() +
   theme_minimal() +
   xlab("Real prevalence observed") +
   ylab("Clinical incidence (annual)") +
   ggtitle("Prevalence - clinical incidence function")
```

### Model specification using greta

To specify the model in greta, first we define the model hyperparameters with priors, then we create any random effects (here the spatiotemporal random effects `epsilon`) and other model objects that depend on them. Then we can define the observation model - linking our generative model to data. 

First we will define the hyperparameters and spatio-temporal process structure and for the spatio-temporal random effects epsilon. We provide two options: the block-circulant embedding approach, and the sparse GP approach, and we demonstrate how to set up objects for both.

We can use the `epiwave.mapping::deifne_bcb_setup()` function to define the objects required for the block-circulant basis approach to simulating IID Gaussian processes over space, and converting them to spatio-temporal Gaussian processes: 

```{r define_bcb}
library(epiwave.mapping)
bcb_setup <- define_bcb_setup(grid_raster)
```

For the sparse GP approach, we instead need to construct a set of spatial 'inducing points' or 'knots' across the region of interest. The number of inducing points controls the accuracy of the approximation - with one inducing point per pixel, this is no longer an approximation. An insufficient number of inducing points will bias inference towards smoother Gaussian processes, though trial and error is required to ensure the inducing point scheme is sufficient to accurately capture the underlying process. For now, we will use a small number of points. We generate points evenly across space in a somewhat ad-hoc way using kmeans clustering. Note that improved efficiency of the approximation could be achieved by weighting inducing points towards more populous areas.

```{r define_inducing}
n_inducing <- 30
pts_sim <- terra::spatSample(grid_raster,
                             n_inducing * 1e3,
                             replace = TRUE,
                             xy = TRUE,
                             values = FALSE,
                             na.rm = TRUE)
                  
inducing <- kmeans(pts_sim, n_inducing)$centers
```

We can plot these to see where they fall and check their coverage:

```{r plot_inducing}
# plot the health facility locations
library(tidyterra)
library(ggplot2)
ggplot() +
  geom_spatraster(data = pop_u5) +
  scale_fill_gradient(
    low = grey(0.9),
    high = grey(0.6),
    transform = "log1p",
    na.value = "transparent",
    guide = "none") +
  geom_point(
    aes(
      y = y,
      x = x
    ),
    data = inducing,
    shape = 16
  ) +
  xlab("") +
  ylab("") +
  theme_minimal() +
  ggtitle("Inducing point locations")
```

Now we can define a space-time Gaussian process over `epsilon`, by creating an isotropic spatial kernel object using the greta.gp R package, a parameter for the AR(1) temporal kernel parameter, and a marginal variance parameter, and passing these to the `epiwave.mapping` functions `sim_gp_bcb()` (for the block-circulant embedding) or `sim_gp_sparse()`.

We first need to define priors for the three hyperparameters of the spatio-temporal process. We define a penalised complexity prior (half-normal) for the marginal variance parameter `sigma`, which shrinks the degree of spatio-temporal correlation towards 0. Since we believe negative residual correlation between time periods (oscillation between negative and positive on a monthly timesteps) to be incredibly unlikely, we define a standard uniform prior on the temporal correlation parameter `theta` so that is positive, but otherwise not informative (the spatial pattern could be either constant, or constantly changing). For the spatial range parameter `phi`, we need to take more care to consider reasonable values in terms of the units of the map (metres). We set up an informative prior to control both the lower and the upper tail of the `phi` - making very short and very long lengthscales (very short-range and long-range spatial correlations) *a priori* unlikely. This helps to resolve the various types of non-identifiability inherent in model-based geostatistics: between the lengthscale and marginal variance parameters, between shorter lengthscales and the spatial covariate parameters `beta` (spatial confounding), and between longer lengthscales and the intercept parameter `alpha`. We do so, by having the user specify upper and lower bounds on the degree of spatial correlation that would be plausible, in terms of the spatial extent of the country under consideration.

```{r spatial_priors}
library(greta)
library(greta.gp)

# Variance of space-time process shrunk towards 0
sigma <- normal(0, 1, truncation = c(0, Inf))
# Temporal correlation forced to be positive, but equally as likely to be strong
# as weak
theta <- uniform(0, 1)

# get the maximum spatial extent of each dimension of the raster, in metres
grid_extent <- as.vector(ext(grid_raster))
x_range <- grid_extent["xmax"] - grid_extent["xmin"]
y_range <- grid_extent["ymax"] - grid_extent["ymin"]
max_distance <- max(x_range, y_range)

# First, find a lower bound for the lengthscale parameter phi, a reasonable
# maximum level of complexity, such that spatial correlation would be 0.1 at
# some fraction of the maximum distance, under a Matern 5/2 covariance
# structure:
phi_lower <- matern_lengthscale_from_correl(distance = max_distance / 5,
                                            correlation = 0.1,
                                            nu = 5/2)

# Do the same for an upper bound. A reasonable spatial correlation for
# some longer distance.
phi_upper <- matern_lengthscale_from_correl(distance = max_distance / 2,
                                            correlation = 0.1,
                                            nu = 5/2)

# Now define a gamma prior on the lengthscale parameter phi, such that the mass
# is low on lengthscales below the lower bound (very complex spatial effects)
# and low on lengthscales above the upper bound (completely flat spatial
# effects). Specifically, we find gamma parameters such that there is only a 10%
# chance of *more complex* spatial correlations than the one above. Note that
# whilst an inverse gamma has the inherent property that it has no mass at 0,
# that is taken care of here by our lower bound, and the gamma is easier to
# solve for these constraints.
gamma_prior_params <- find_gamma_parameters(upper_value = phi_upper,
                                            upper_prob = 0.1,
                                            lower_value = phi_lower,
                                            lower_prob = 0.1)

# # we can double check this converged OK:
# round(pgamma(phi_lower,
#              shape = gamma_prior_params$shape,
#              rate = gamma_prior_params$rate), 2)
# round(1 - pgamma(phi_upper,
#                  shape = gamma_prior_params$shape,
#                  rate = gamma_prior_params$rate), 2)

# define the prior over the lengthscale parameter phi
phi <- gamma(shape = gamma_prior_params$shape,
             rate = gamma_prior_params$rate)

```

To use the block-circulant approach we need to define a spatial kernel in greta.gp, with only a single lengthscale parameter, and pass this to `epiwave.mapping::sim_gp_bcb()` along with our setup for the BCB approach. I've commented this out (as well as not evaluating the block), since running this block in addition. to the sparse approach we use below will make the model run very slowly.

We can construct the space-time process using a sparse GP with `epiwave.mapping::sim_gp_sparse()`. In this case (*for reasons*), to specify an isotropic spatial kernel we need to pass the lengthscale parameter to the `greta.gp` kernel function twice, once for each dimension:
```{r define_epsilon_sparse}
# note we pass the lengthscale argument twice
k_space <- mat52(lengthscales = c(phi, phi),
                 variance = 1)

# set time periods
start_year <- 2021
n_years <- 3

epsilon <- epiwave.mapping::sim_gp_sparse(
  inducing_points = inducing,
  grid_raster = grid_raster,
  n_times = n_years * 12,
  space_kernel = k_space,
  time_correlation = theta,
  sigma = sigma
)
```

We can visually check that our prior is giving us reasonable values of epsilon by doing some prior simulations and then plotting them:
```{r gp_prior_check, fig.width = 10, fig.height=9}
n_sims <- 3
times_plot <- 1:5
prior_sims <- calculate(epsilon,
                        phi,
                        theta,
                        sigma,
                        nsim = n_sims)

# for each simulation, create an empty multilayer raster, fill it in with the
# simulations for each of the first three years, and plot it
plot_list <- list() 
for (sim in seq_len(n_sims)) {
  
  plot_raster <- do.call(c,
                         replicate(length(times_plot),
                                   grid_raster,
                                   simplify = FALSE))
  names(plot_raster) <- paste("timestep", times_plot)
  
  for (i in seq_along(times_plot)) {
    plot_raster[[i]][cells(grid_raster)] <- prior_sims$epsilon[sim, , i]
  }

  plot_list[[sim]] <- ggplot() +
    geom_spatraster(data = plot_raster) +
    scale_fill_distiller(
      palette = "PiYG",
      na.value = "transparent",
      guide = "none") +
    facet_wrap(~lyr, nrow = 1) +
    theme_minimal() +
    xlab("") +
    ylab("") +
    ggtitle(label = paste("Simulation", sim),
            subtitle = sprintf(
              "phi (space) = %s, theta (time) = %s, sigma = %s",
              round(prior_sims$phi[sim, 1, 1]),
              round(prior_sims$theta[sim, 1, 1], 2),
              round(prior_sims$sigma[sim, 1, 1], 2)
            ))
}

library(patchwork)
combined_plot <- plot_list[[1]]
for (sim in 2:n_sims) {
  combined_plot <- combined_plot / plot_list[[sim]]
}
combined_plot
```
You can run that block multiple times to see different realisations, and check they conform to you prior expectation about the spatio-temporal random process.

Now we have the object for the spatio-temporal random effects, we define our remaining hyperparameters for the covariate effects, and reporting fraction.

Here we use an arbitrary set of priors, that happen to be those used to simulate the 'true' parameters in `epiwave.mapping::sim_data()`. In a practical application, the user should give these a lot more thought. 
```{r define_and_priors}
# intercept for logit-sacle daily infection incidence
alpha <- normal(log(1e-4), 1) #Have a think about this parameter
# slope for fixed-effects terms (static covariayes) on daily infection incidence
beta <- normal(0, 1, dim = nlyr(covariates_raster))

# Time-varying covariate effects (each over 36 months)
beta_dynamic        <- normal(0, 1, dim = c(nlyr(EVI_covariates_raster), nlyr(EVI_covariates_raster)))

# ITN covariate effects (non-negative)
beta_itn        <- normal(0, 1, dim = c(ncol(ITN_temporal_cov_36),ncol(ITN_temporal_cov_36)), truncation = c(0, Inf))

prior_hist <- function(x, title = "Histogram of Prior"){
   sims <- calculate(x, nsim = 1000)
   hist(sims$x,main=title, breaks = 100)
 }
# fraction of all clinical cases reported in case counts
#r <- beta(10, 5)
library(bayesplot)

prior_hist(sigma, "Histogram of sigma")
prior_hist(theta, "Histogram of theta")
prior_hist(alpha, "histogram of alpha")
prior_hist(beta, "histogram of beta-static")
prior_hist(beta_dynamic, "histogram of beta_dynamic")
prior_hist(beta_itn, "histogram of beta_itn")
```

Then we can define the deterministic mapping from these, through infection incidence, to our observed data. First we compute the daily incidence of new infections, per cell, per timestep.
```{r Static_covs_effect}
#Static component (constant across months)
# matrix-multiply with beta, and add alpha, to get the spatial fixed 
static_ef <- alpha + X %*% beta
```



1. We check the dynamic covariates effect
Use a beta after summing covariates
```{r dynamic_covs_check, fig.width = 10, fig.height=9}
n_sims <- 3
times_plot <- 1:36

# Calculate the dynamic effect matrix
dynamic_ef_mat <- do.call(cbind, lapply(1:n_months, function(i) {
  (EVI_temporal_cov[, i] +
   LST_day_temporal_cov[, i] +
   LST_night_temporal_cov[, i] +
   CHIRPS_temporal_cov[, i] +
   TCB_temporal_cov[, i] +
   TCW_temporal_cov[, i]) %*% beta_dynamic[i]
}))

# Run simulations
dynamic_sims <- greta::calculate(dynamic_ef_mat, nsim = n_sims)

# Select a single simulation (e.g., the first)
sim <- 1

# Create multi-layer raster for all 36 months
plot_raster <- do.call(c,
                       replicate(length(times_plot), grid_raster, simplify = FALSE))
names(plot_raster) <- paste("Month", times_plot)

for (i in seq_along(times_plot)) {
  plot_raster[[i]][cells(grid_raster)] <- dynamic_sims$dynamic_ef_mat[sim, , i]
}

# Plot all 36 months in a 6x6 grid
plot_dynamic_covariates <- ggplot() +
  geom_spatraster(data = plot_raster) +
  scale_fill_distiller(
    palette = "PiYG",
    na.value = "transparent"
  ) +
  facet_wrap(~lyr, ncol = 6, nrow = 6) +
  theme_minimal() +
  xlab("") +
  ylab("") +
  ggtitle("Dynamic covariate additive effect across 36 months",
          subtitle = sprintf("Simulation %d — phi = %.2f, theta = %.2f, sigma = %.2f",
                             sim,
                             prior_sims$phi[sim, 1, 1],
                             prior_sims$theta[sim, 1, 1],
                             prior_sims$sigma[sim, 1, 1]))

plot_dynamic_covariates
```
2. Dynamic covariates effect
Use a beta after taking the mean from covariates  
```{r dynamic_covs_check_2, fig.width = 10, fig.height=9}
n_sims <- 3
times_plot <- 1:36

# Calculate the dynamic effect matrix
dynamic_ef_mat <- do.call(cbind, lapply(1:n_months, function(i) {
  rowMeans(cbind(
    EVI_temporal_cov[, i],
    LST_day_temporal_cov[, i],
    LST_night_temporal_cov[, i],
    CHIRPS_temporal_cov[, i],
    TCB_temporal_cov[, i],
    TCW_temporal_cov[, i]
  )) %*% beta_dynamic[i]
}))
# Run simulations
dynamic_sims <- greta::calculate(dynamic_ef_mat, nsim = n_sims)
# Select a single simulation (e.g., the first)
sim <- 1
# Create multi-layer raster for all 36 months
plot_raster <- do.call(c,
                       replicate(length(times_plot), grid_raster, simplify = FALSE))
names(plot_raster) <- paste("Month", times_plot)

for (i in seq_along(times_plot)) {
  plot_raster[[i]][cells(grid_raster)] <- dynamic_sims$dynamic_ef_mat[sim, , i]
}

# Plot all 36 months in a 6x6 grid
plot_dynamic_covariates <- ggplot() +
  geom_spatraster(data = plot_raster) +
  scale_fill_distiller(
    palette = "PiYG",
    na.value = "transparent"
  ) +
  facet_wrap(~lyr, ncol = 6, nrow = 6) +
  theme_minimal() +
  xlab("") +
  ylab("") +
  ggtitle("Dynamic covariate mean effect across 36 months",
          subtitle = sprintf("Simulation %d — phi = %.2f, theta = %.2f, sigma = %.2f",
                             sim,
                             prior_sims$phi[sim, 1, 1],
                             prior_sims$theta[sim, 1, 1],
                             prior_sims$sigma[sim, 1, 1]))

plot_dynamic_covariates
```
3. Dynamic covaraites effect
Use individual betas per covariate (preferred if you want distinct effects)
```{r dynamic_covs_check_3, fig.width = 10, fig.height=9}
n_sims <- 3
times_plot <- 1:36
beta_EVI <- normal(0,1, dim = nlyr(EVI_covariates_raster))
beta_LST_day <- normal(0,1, dim = nlyr(EVI_covariates_raster))
beta_LST_night <- normal(0,1, dim = nlyr(EVI_covariates_raster))
beta_CHIRPS <- normal(0,1, dim = nlyr(EVI_covariates_raster))
beta_TCB <- normal(0,1, dim = nlyr(EVI_covariates_raster))
beta_TCW <- normal(0,1, dim = nlyr(EVI_covariates_raster))
# Calculate the dynamic effect matrix
dynamic_ef_mat <- do.call(cbind, lapply(1:n_months, function(i) {
  EVI_temporal_cov[, i]        * beta_EVI[i] +
  LST_day_temporal_cov[, i]    * beta_LST_day[i] +
  LST_night_temporal_cov[, i]  * beta_LST_night[i] +
  CHIRPS_temporal_cov[, i]     * beta_CHIRPS[i] +
  TCB_temporal_cov[, i]        * beta_TCB[i] +
  TCW_temporal_cov[, i]        * beta_TCW[i]
}))
# Run simulations
dynamic_sims <- greta::calculate(dynamic_ef_mat, nsim = n_sims)
# Select a single simulation (e.g., the first)
sim <- 1
# Create multi-layer raster for all 36 months
plot_raster <- do.call(c,
                       replicate(length(times_plot), grid_raster, simplify = FALSE))
names(plot_raster) <- paste("Month", times_plot)

for (i in seq_along(times_plot)) {
  plot_raster[[i]][cells(grid_raster)] <- dynamic_sims$dynamic_ef_mat[sim, , i]
}
# Plot all 36 months in a 6x6 grid
plot_dynamic_covariates <- ggplot() +
  geom_spatraster(data = plot_raster) +
  scale_fill_distiller(
    palette = "PiYG",
    na.value = "transparent"
  ) +
  facet_wrap(~lyr, ncol = 6, nrow = 6) +
  theme_minimal() +
  xlab("") +
  ylab("") +
  ggtitle("Dynamic covariate distinct effect across 36 months",
          subtitle = sprintf("Simulation %d — phi = %.2f, theta = %.2f, sigma = %.2f",
                             sim,
                             prior_sims$phi[sim, 1, 1],
                             prior_sims$theta[sim, 1, 1],
                             prior_sims$sigma[sim, 1, 1]))

plot_dynamic_covariates
```
4. Dynamic covaraites effect
Use individual betas per covariate and take the mean of effects
```{r dynamic_covs_check_4, fig.width = 10, fig.height=9}
n_sims <- 3
times_plot <- 1:36
beta_EVI <- normal(0,1, dim = nlyr(EVI_covariates_raster))
beta_LST_day <- normal(0,1, dim = nlyr(EVI_covariates_raster))
beta_LST_night <- normal(0,1, dim = nlyr(EVI_covariates_raster))
beta_CHIRPS <- normal(0,1, dim = nlyr(EVI_covariates_raster))
beta_TCB <- normal(0,1, dim = nlyr(EVI_covariates_raster))
beta_TCW <- normal(0,1, dim = nlyr(EVI_covariates_raster))
# Calculate the dynamic effect matrix
dynamic_ef_mat <- do.call(cbind, lapply(1:n_months, function(i) {
  (EVI_temporal_cov[, i]        * beta_EVI[i] +
  LST_day_temporal_cov[, i]    * beta_LST_day[i] +
  LST_night_temporal_cov[, i]  * beta_LST_night[i] +
  CHIRPS_temporal_cov[, i]     * beta_CHIRPS[i] +
  TCB_temporal_cov[, i]        * beta_TCB[i] +
  TCW_temporal_cov[, i]        * beta_TCW[i])  / 6  # divide to get the mean
}))
# Run simulations
dynamic_sims <- greta::calculate(dynamic_ef_mat, nsim = n_sims)
# Select a single simulation (e.g., the first)
sim <- 1
# Create multi-layer raster for all 36 months
plot_raster <- do.call(c,
                       replicate(length(times_plot), grid_raster, simplify = FALSE))
names(plot_raster) <- paste("Month", times_plot)

for (i in seq_along(times_plot)) {
  plot_raster[[i]][cells(grid_raster)] <- dynamic_sims$dynamic_ef_mat[sim, , i]
}
# Plot all 36 months in a 6x6 grid
plot_dynamic_covariates <- ggplot() +
  geom_spatraster(data = plot_raster) +
  scale_fill_distiller(
    palette = "PiYG",
    na.value = "transparent"
  ) +
  facet_wrap(~lyr, ncol = 6, nrow = 6) +
  theme_minimal() +
  xlab("") +
  ylab("") +
  ggtitle("Dynamic covariate mean distinct effect across 36 months",
          subtitle = sprintf("Simulation %d — phi = %.2f, theta = %.2f, sigma = %.2f",
                             sim,
                             prior_sims$phi[sim, 1, 1],
                             prior_sims$theta[sim, 1, 1],
                             prior_sims$sigma[sim, 1, 1]))

plot_dynamic_covariates
```


5. Dynamic ITN use effect
```{r dynamic_covs_check_5, fig.width = 10, fig.height=9}
n_sims <- 3
times_plot <- 1:36

# Calculate the dynamic effect matrix
dynamic_ef_mat <- ITN_temporal_cov_36 %*% beta_itn
# Run simulations
dynamic_sims <- greta::calculate(dynamic_ef_mat, nsim = n_sims)
# Select a single simulation (e.g., the first)
sim <- 1
# Create multi-layer raster for all 36 months
plot_raster <- do.call(c,
                       replicate(length(times_plot), grid_raster, simplify = FALSE))
names(plot_raster) <- paste("Month", times_plot)

for (i in seq_along(times_plot)) {
  plot_raster[[i]][cells(grid_raster)] <- dynamic_sims$dynamic_ef_mat[sim, , i]
}
# Plot all 36 months in a 6x6 grid
plot_dynamic_covariates <- ggplot() +
  geom_spatraster(data = plot_raster) +
  scale_fill_distiller(
    palette = "PiYG",
    na.value = "transparent"
  ) +
  facet_wrap(~lyr, ncol = 6, nrow = 6) +
  theme_minimal() +
  xlab("") +
  ylab("") +
  ggtitle("Dynamic ITN use effect across 36 months",
          subtitle = sprintf("Simulation %d — phi = %.2f, theta = %.2f, sigma = %.2f",
                             sim,
                             prior_sims$phi[sim, 1, 1],
                             prior_sims$theta[sim, 1, 1],
                             prior_sims$sigma[sim, 1, 1]))

plot_dynamic_covariates
```