---
output:
  github_document:
    toc: true
    toc_depth: 3
    pandoc_args: --webtex
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = TRUE,
  comment = "#>",
  message = FALSE
)
```

# epiwave.mapping

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Prototype R package and example code for computationally efficient, fully Bayesian semimechanistic spatio-temporal mapping of disease infection incidence simultaneously from spatially-aggregated (e.g. at health facility level) clinical case count timeseries, and targeted infection prevalence surveys at specific point locations. This model and software is based on ideas in the epiwave R package and uses some functions implemented there, but it currently relies on the user to develop their own model using the greta R package, rather than providing wrapper functions to set up the model for the user.

## Installation

You can install the package from github, using the `remotes` package:

```{r install, eval = FALSE}
remotes::install_github("idem-lab/epiwave.mapping",
                        dependencies = TRUE)
```

## Model

Our main aim is to model, and generate spatio-temporal predictive maps of, variation in *infection incidence*: the average number of new infections per time period, per member of the population. We model the logarithm of infection incidence with a space-time Gaussian process. Since infection incidence is rarely perfectly observed, we infer it from two independent and complementary datastreams:

- **Clinical case timeseries**: Counts of the numbers of cases reported over time, but potentially spatially aggregated across e.g. the catchment areas of health facilities, or the entire region, rather than at spatially precise coordinates

- **Infection prevalence survey points**: Random samples of individuals at a specific point location, with the number tested and number positive recorded. Typically these data are available only very infrequently.

Since neither datastream provides a direct estimate of infection incidence, we model the generative observation process yielding each datastream, using informative priors for key parameters wherever possible. In order to model the full spatial process generating these data, the model requires estimation of the infection incidence across every pixel in a raster grid covering the region of interest - incidence across all cells is aggregated-up to compute the catchment-level case counts. In this way, the model fully accounts for spatial variation in environmental conditions (represented by regression against environmental predictors) and in the residual space-time process across all cells. This model structure leads to a non-linear and non-factorising likelihood which negates many of the computational tricks used in the INLA software for fitting geostatistical models that can be structured as Gaussian Markov random fields. Instead we employ fully Bayesian inference using Hamiltonian Monte Carlo (providing significant flexibility in model specification) and computational approximations to evaluating the Gaussian process across the entire spatial extent, of which a number of options are available.

Below illustrate two methods suited to mapping a separable space-time Gaussian process: simulation with a block-circulant embedding, and with a sparse Gaussian process approximation. The former is similar to the approach used in the `lgcp` R package, and detailed in Diggle et al. (2013). We note that the clinical incidence observation model we employ is a particular case of a Log-Gaussian Cox process. Our contribution is accounting for delays in reporting, and linking model this to prevalence survey data, in a practical and easily-extensible framework.

### Clinical incidence model 

We model the observed number of clinical cases $C_{h,t}$ of our disease of interest in health facility $h$ during discrete time period $t$ as a Poisson random variable:

$$
C_{h,t} \sim \text{Poisson}(\hat{C}_{h,t})
$$
The expectation of this Poisson random variable (the modelled/expected number of clinical cases) is given by a weighted sum of (unobserved but modelled) expected pixel-level clinical case counts $\hat{C}_{l,t}$ at each of $L$ pixel locations $l$:

$$
\hat{C}_{h,t} = \sum_{l=1}^{L}{\hat{C}_{l,t}w_{l,h}}
$$

where weights $w_{l,h}$ give the 'membership' of the population in each pixel location to each health facility, where each case at a given location $l$ has probability $w_{l,h}$ of reporting at health facility $h$, and therefore $\sum_{l=1}^L w_{l,h} = 1$. In practice this could be either a proportional (fractions of the population attend different health facilities) or a discrete (the population of each pixel location attends only one nearby health facility) mapping.

At each pixel location, we model the unobserved clinical case count $\hat{C}_{l,t}$ from the modelled number of new infections $\hat{I}_{l,t'}$ at the same location, during previous time periods $t'$, from the fraction of infections $\gamma_{l,t'}$ at that location and previous time period that would result in a recorded clinical case, and a probability distribution $\pi(t-t')$ over delays $t-t'$ from infection to diagnosis and reporting:

$$
\hat{C}_{l,t} = \sum_{t-t' = 0}^{\tau_\pi}{\hat{I}_{l,t'} \, \gamma_{l,t'} \, \pi(t-t')}
$$
The distribution $\pi(\Delta_t)$ gives the discrete and finite (support on $\Delta_t \in (0, \tau_\pi)$) probability distribution over delays from infection to reporting, indexed on the time-periods considered in the model. This temporal reweighting to account for a distribution over possible delays can be considered as a 'discrete-time convolution' with $\pi(\Delta_t)$ the 'kernel'. Below we discuss efficient methods for computing these convolutions in greta.

### Infection prevalence model 

We model the observed number of individuals who test positive for infection $N^+_{l,t}$ in an infection prevalence survey at location $l$ at time $t$ as a binomial sample, given the number of individuals tested $N_{l,t}$, and the modelled prevalence of infections $\hat{p}_{l,t}$ across the population at that time/place:

$$
N^+_{l,t} \sim \text{Binomial}(N_{l,t}, \hat{p}_{l,t})
$$
Similarly to clinical incidence, we model the infection prevalence at a given location and time as a discrete-time convolution over previous infection counts, divided by the total population in the pixel $M_l$, and the duration of the time-period in days $d$:

$$
\hat{p}_{l,t} = \frac{1}{M_l}\sum_{t-t' = 0}^{\tau_q}{q(t-t') \, \hat{I}_{l,t'} \, d^{-1}}
$$

In this case, the kernel $q(t-t')$ is not a probability distribution, but the kernel of a convolution operation that maps the daily average number of new infections in previous time periods $t'$ to the number of individuals who would test positive (to the diagnostic used) on a randomly-selected day in the timeperiod `t`. This kernel, which is specific to the tim eperiod chosen for modelling the temporal process, can be computed from a related function: $q_{\text{daily}}(\Delta_t)$ which gives the probability that an individual will test positive $\Delta_t$ days after infection. The integral of this function is the average number of days an infected person would test positive for. The function $q_{\text{daily}}()$ can be estimated from empirical data on how the test sensitivity and duration of infection vary over time since infection. This package provides functionality: `transform_convolution_kernel()` to construct timeperiod-specific kernels such as $q()$ from daily kernels like $q_{\text{daily}}()$ for an arbitrary modelling timeperiod.

By convolving of the number of new infections *per day* $\hat{I}_{l,t'} \, d^{-1}$ in all previous time periods with the fraction of those that would test positive in a survey in time period $t$, gives an estimate of the *number* of positive-testing people in the population at location $l$, on any given day within the time period $t$. Dividing this by the population of that location, $M_l$, therefore yields an estimate of the population proportion testing positive on any given day; the parameter of the binomial distribution that captures the prevalence survey data.

Note that our definition of $\hat{p}_{l,t}$ is the population prevalence of infections *that would test positive using that diagnostic method* rather than the true fraction infected/infectious at any one time. We also assume here that tests have perfect specificity, though the model can easily be adapted to situations where that is not the case.

### Infection incidence model 

The expected number of new infections $\hat{I}_{l,t'}$ in location $l$ during time period $t$ is modelled as the product of the population $M_l$ at that location, the length of the timeperiod in days $d$, and the daily infection incidence $f_{l,t}$ at that location and time:

$$
\hat{I}_{l,t'} = d \, M_l \, \hat{f}_{l,t}
$$

Whilst we mechanistically model the observation processes yielding our data types, we employ a geostatistical approach to modelling spatio-temporal variation in infection incidence, with spatio-temporal covariates $\mathbf{X}_{l,t}$ and a space-time random effect $\epsilon_{l,t}$ with zero-mean Gaussian process prior:

$$
\text{log}(f_{l,t}) = \alpha +\mathbf{X}_{l,t} \beta + \epsilon_{l,t}
$$

$$
\epsilon \sim GP(0, \mathbf{K})
$$

where $\alpha$ is a scalar intercept term, $\beta$ is a vector of regression coefficients against the environmental covariates, and $\mathbf{K}$ is the space-time covariance function of the Gaussian process over random effects $\epsilon_{l,t}$.

There are many choices of space-time covariance structure for $\mathbf{K}$, though we use a separable combination of an isotropic spatial covariance function with a temporal covariance function, to enable the use of a range of computationally efficient simulation and calculation methods:

$$
K_{l,t,l',t'} = \sigma^2 \, K_{\text{space}}(||l-l'||; \phi) \, K_{\text{time}}(|t-t'|; \theta)
$$
where $\sigma^2$ is the marginal variance (amplitude) of the resultant Gaussian process, $K_{\text{space}}(.; \phi)$ is an (unit variance) isotropic spatial kernel on euclidean distances $||l-l'||$ with parameter $\phi > 0$ controlling the range of spatial correlation, and $K_{\text{time}}(|t-t'|; \theta)$ is a a temporal kernel on time differences $|t-t'|$, with parameter $\theta$ controlling the range of temporal correlation. Again for computational reasons, we prefer Markovian kernels for $K_{\text{time}}$.

### Infection prevalence - clinical incidence relationship

As described above, the spatially- and temporally-varying parameter $\gamma_{l,t}$  gives the fraction of new infections that would result in a recorded clinical case. In the absence of additional data to that described above, it is unlikely that spatio-temporal variation in this parameter can be inferred. For some diseases - e.g. those where previous infections confer little immune protection from symptoms of future infections - it may be sufficient to model this as a constant, in which case it can be inferred statistically. However for many diseases, there is likely to be a non-linear relationship between the infection incidence in a given population, and the fraction of those infections that result in a clinical case. Where repeat infections decrease the clinical severity of subsequent infections, $\gamma_{l,t}$ is likely to manifest as a monotone decreasing function of infection incidence $f_{l,t}$. In this scenario, it would be preferable to account for this relationship using external information. In the case of malaria, this relationship is often explored using empirical data in terms of the relationship between infection prevalence and clinical incidence. Where a previously-estimated relationship between infection prevalence and clinical incidence is available, the relationship between infection incidence and clinical incidence can be inferred as follows.

We assume that a pre-determined function $g()$ is available that maps from the 'true' population infection prevalence $p_{i}$ to the 'true' underlying clinical incidence $c_{i}$ in some population and setting $i$, for which the corresponding infection incidence is held constant:

$$
c_{i} = g(p_{i})
$$

In this case, $c_{i}$ gives the incidence of all clinical episodes regardless whether those cases would be recorded in a health system. We therefore assume that $g()$ is estimated against empirical estimates of $p_{i}$ and of $c_{i}$ that would be observed under some intensive surveillance (e.g. a prospective study), or other high-quality source of information. We therefore capture imperfect reporting in the health system under which observed case counts $C_{i}$ are collected via some constant reporting rate $r$:

$$
C_{i} = r \, d \, M_i \, c_{i}
$$
where $d$ is the duration of time periods over which the counts are made, and $M_i$ is the size of this population (converting from clinical incidence to the number of clinical cases, as above). Under the assumption (from $g()$) that infection incidence $f_{i}$ is constant over the period to which $c_{i}$ and $p_{i}$ correspond, these two quantities can therefore be related as:

$$
c_{i} = \kappa_i f_{i}
$$
where $\kappa_i$ is the fraction of infections at location $i$ resulting in a clinical case (recorded or otherwise), and:
$$
p_{i} = q^* \, f_{i}
$$
where $q^*$ is the average number of subsequent days on which each new infection would test positive, if tested once per day (following from the convolution above, in the case that $f_{l,t}$ and therefore $I_{l,t}$ are constant over time); the integral of the detectability function $q()$: 
$$
q^* = \int_{0}^{\tau_q}{q(t) dt}
$$
This scalar value $q^*$ can equally be interpretted as the average number of previous days worth of infections that are detected on the day of a prevalence survey. In practice, $q^*$ can be estimated by a sufficiently-fine resolution discrete sum over $q(t)$.

Combining the above, we have that:

$$
c_i = \kappa_i \, f_i = g(q^* f_i)
$$
and therefore:
$$
\kappa_i = g(q^* f_i) / f_i
$$
simce $\gamma_i = r \, \kappa_i$, and applying this relationship to our spatio-temporally varying infection incidences $f_{l,t}$, we can compute $\gamma_{l,t}$ as:
$$
\gamma_{l,t} = r \, g(q^* f_{l,t}) \, / \, f_{l,t}
$$

where $q^*$ is a scalar constant that can be computed *a priori* from $q$, $g()$ is a deterministic function, also known *a priori*, $f_{l,t}$ is our inference target, the daily infection incidence, and $r$ is the only remaining free parameter in this relationship, which is identifiable in the model.

In practice, the functional form of $g()$ may well permit the computational complexity of this term to be reduced, either by computing a simpler analytic solution, or approximating it with a functionally simpler form with near-equivalent shape. This will likely aid in reducing the computational cost of fitting the model.

## Computational approaches

### Inference

We fit the model via fully Bayesian inference, using MCMC (Hamiltonian Monte Carlo) in the greta R package. Whilst a number of alternative inference algorithms have been proposed and widely adopted for Bayesian and non-Bayesian estimation of spatio-temporal Gaussian process models (e.g. INLA), in out case, the aggregation of clinical cases across catchments (a feature required by the model structure) leads to a non-factorising likelihood, which nullifies many of the computational advantages of that method. Further, the mapping from the Gaussian process to the expectations of the two sampling distributions is non-linear (due to sums of exponents in various places), requiring computationally costly linearisation of the non-linearities. By employing a fully-Bayesian MCMC approach, and focussing on computationally-efficient model and implementation choices, rather than approximations, we are able to estimate the model parameters relatively quickly, with full treatment, propagation, and quantification of model uncertainty, and retaining the ability to easily modify and interrogate the model.

### Discrete-time convolution

This discrete convolutions (weighted sums over previous time points) used to compute $\hat{C}_{l,t}$ and $\hat{p}_{l,t}$ can be computationally intensive, depending on the number of time periods modelled. A number of efficient computational approaches exist to overcome these, and the optimal approach to use in greta depends on the size of $\tau^{max}$: the maximum duration of the delay in terms of the number of timeperiods being considered. Where $\tau^{max}$ is relatively large (say, $\tau^{max} > 3$), implementation of the convolution as a matrix multiply is likely to be the most efficient in greta. If $\tau^{max} > 3$ and the number of time periods being modelled is much larger than $\tau^{max}$, implementing this as a sparse matrix multiply (skipping computation on zero elements of the convolution matrix) is likely to be most efficient. Where $\tau^{max}$ is small (say, $\tau^{max} \leq 3$) the convolution can instead be computed with a sum of dense vectorised additions and subtractions. These convolution approaches are implemented in this package, and demonstrated below.

### Gaussian process simulation

Naive implementation of the full Gaussian process (GP) model is typically very computationally intensive, due to the fact that the most expensive step (inverting a dense covariance matrix) scales cubically with the number of evaluation points. In the model we propose (and as in a Log-Gaussian Cox process), we need to evaluate the GP at every pixel location and time period in the study frame, so that we can aggregate the expected clinical case count to compute the likelihood. This results in a computationally impractical algorithm for all but the smallest study frames. Some form of computational approximation is required.

One straightforward approximation would be to employ a full (non-approximate) GP, but to evaluate the combined process (GP and fixed effects regression) at only a subset of locations and times in the study frame and then to approximate the health-facility level clinical case count calculation (a set of integrals over space) with some smaller finite sum. Whilst general, this would also result in an approximation to the fixed-effects environmental regression part of the model. While the GP could feasibly be smooth and well approximated in this way, the environmental covariates are likely to be more complex and thus potentially quite poorly approximated by a finite sum.

An alternative approach would be to evaluate the GP and the fixed effects at all pixels, but to replace the full GP with an approximation. Candidate GP approximation approaches include the SPDE approximation to a Matern-type spatial kernel, on a computational mesh (as used in INLA, [Lindgren *et al.*, 2011](https://doi.org/10.1111/j.1467-9868.2011.00777.x)), a sparse GP method over a limited set of inducing points ([Quinonera-Candela & Rasmussen, 2005](https://jmlr.org/papers/v6/quinonero-candela05a.html) - see `greta.gp` and note the similarity to a Guassian predictive process per [Banerjee *et al.*, 2009](https://doi.org/10.1111/j.1467-9868.2008.00663.x)), or one of the closely-related penalised spline methods (see `greta.gam`). Implementation of the SPDE approximation in an HMC setting requires implementation of sparse matrix methods adapted to automatic differentiation ([Durrande *et al.* 2019](https://doi.org/10.48550/arXiv.1902.10078), greta version in development), but would likely be preferable where the underlying Gaussian process has short lengthscale. In the case of a comparatively smooth underlying GP, the sparse o spline-based methods are likely to be a better fit, since they have lower computational complexity in this case. Given the spatially-aggregated observation model, it seems likely that only comparatively smooth GP realisations could be identified in this model, and therefore favour sparse or spline methods. 

An appealing third alternative in this case is to compute the full spatial GP for every pixel in a regular grid (the same we use to record spatial covariate values), by exploiting the block-circulant structure in the resulting spatial covariance matrix. This requires using projected coordinates, and expanding the spatial study frame (by a factor of 2 in each dimension) to map the projection onto a torus in such a way that the distances between pairs of pixels are preserved. This approach enables simulation of the GP across all pixels, for a given time period, via the fast fourier transform (FFT) - which scales only linearly with the number of locations considered (? check when have wifi). This enables us to simulate values of the spatial GP across all pixels very cheaply, with no need to approximate the GP. In separable combination with a Markovian temporal kernel, this enables very rapid inference. However the very large number of random variables required to be sampled (at least four times as many as pixels and time points under study) leads to a very large parameter space for sampling, and can slow down efficient MCMC sampling considerably.

## Example application

We demonstrate the model with application to mapping the (simulated) infection incidence of malaria in Kenya from simulated clinical incidence and infection prevalence data. Using the `sim_data()` function from this package, we use a real grid of environmental covariates, but simulate the locations of health facilities, prevalence survey locations, and all datasets to which we will then fit the model.


Define reference image, population and admin masks ----
```{r}
library(raster)
library(terra)
library(sf)
# Country-level shapefiles
dir_path <- "/mnt/s3/stratification/CountryFolders/Nigeria"
# Raster locations
raster_dir <- paste0(dir_path,"/Common files/Covariates/1km/")

# Country boundries
shape <- st_read(file.path(dir_path,"Common files/Shapefiles/NGA_LGAs.shp")) 

# Population under 5 rasterfile cropped and masked
pop_u5 <- raster("/mnt/s3/stratification/CountryFolders/Nigeria/Common files/Population/u5_grid3_1km_new.tif")%>% 
  crop(shape) %>% mask(shape)

# Change projection to UTM
new_crs <-"epsg:26332"  #"epsg:21036"
pop_u5 <- terra::project(rast(pop_u5), new_crs)

# Ensure that the 'shape' is a valid sf object
# Make sure 'shape' is a valid Simple Features object (sf)
shape <- st_make_valid(shape)  # Fix invalid geometries

# Perform the union operation after ensuring the geometries are valid
admin0 <- st_union(shape)
plot(admin0)
NGA<-admin0

# Assuming 'pop_u5' is a SpatRaster object
# library(terra)
# coordinates_pop_u5 <- xyFromCell(pop_u5, 1:ncell(pop_u5))  # Extract coordinates from 'pop_u5' SpatRaster
# 
# # Assuming 'pop_u5' is a SpatRaster object and 'admin0' is an sf object
# # Extract the coordinates from the admin0 polygon using sf methods
# admin0_coords <- st_coordinates(admin0)  # Extract coordinates from admin0 (sf object)
# 
# # Extract the x and y coordinates of the polygon geometry
# admin0_x <- admin0_coords[, 1]
# admin0_y <- admin0_coords[, 2]
# 
# # Now, use the point.in.polygon function (from the 'sp' package)
# in.country <- point.in.polygon(
#   coordinates_pop_u5[,1],  # x-coordinates from pop_u5
#   coordinates_pop_u5[,2],  # y-coordinates from pop_u5
#   admin0_x,                # x-coordinates of admin0 polygon
#   admin0_y                 # y-coordinates of admin0 polygon
# )
# 
#in.country <- which(in.country==1 & values(pop_u5)>0)# do not make prediction if population not above 0


# Create an index vector of each pixel, along with their coordinates
population   <- values(pop_u5)
valid.pix     <- which(!is.na(population) & population>0)
             
im.coords     <- xyFromCell(pop_u5, valid.pix)
valid.coords      <- im.coords
```

```{r project_rasters}
library(terra)
# define a template raster representing our compute grid: 0 in land cells, NA
# elsewhere
grid_raster <- pop_u5 * 0
names(grid_raster) <- NULL

# Pick projection as UTM zone 36S (Uganda, Kenya, Tanzania)
#Nigeria is covered by three UTM zones, namely: Zones 31, 32 and 33
new_crs <-"epsg:26332"  #"epsg:21036"
# project these files
grid_raster <- terra::project(rast(grid_raster), new_crs)
```


```{r}
library(raster)
  library(terra)
library(tidyverse)
dir_path <- "/mnt/s3/stratification/CountryFolders/Nigeria"

# Raster locations
raster_dir <- paste0(dir_path,"/Common files/Covariates/1km/")
# Raster files
covariate.names <- list(
  "accessibility_to_cities_2015_v1.0.tif",
  "Aridity_Index_v2.Synoptic.Overall.Data.1km.Data.tif",
  "distance_to_water.tif",
  "SRTM_Elevation.Synoptic.Overall.Data.1km.mean.tif",
  "EVI_v6.2018.Annual.mean.1km.Data.tif",
  "LST_Day_v6.2018.Annual.mean.1km.Data.tif",
  "LST_Night_v6.2018.Annual.mean.1km.Data.tif",
  "PET_v2.Synoptic.Overall.Data.1km.Data.tif",
  "SRTM_SlopePCT_Corrected.Synoptic.Overall.Data.1km.Data.tif",
  "DMSP.F18_Stable_Lights_v4c.2013.Annual.Data.1km.Data.tif",
  "TCB_v6.2018.Annual.mean.1km.Data.tif",
  "TCW_v6.2018.Annual.mean.1km.Data.tif",
  "TempSuitability.Pf.Index.1k.2010.global.tif",
  "tree_fraction_any_percent.tif",
  "HydroSHEDS_TWI.tif")

# # Labels
# names(covariate.names) <- c(
#   "Accesibility to cities",
#   "Aridity Index",
#   "Distance to water",
#   "Elevation",
#   "EVI",
#   "LST Day",
#   "LST Night",
#   "PET",
#   "Slope",
#   "Stable Lights",
#   "TCB",
#   "TCW",
#   "TSI",
#   "Tree fraction",
#   "TWI")

# Number of covariates
Ncovariates <- length(covariate.names)

# Transformations to apply to 
transformation.types <- c("Exponential",
                          "Normal",
                          "Exponential",
                          "Exponential",
                          "Normal",
                          "Normal",
                          "Normal",
                          "Normal",
                          "Exponential",
                          "Exponential",
                          "Normal",
                          "Normal",
                          "Normal",
                          "Normal",
                          "Normal"
)

# Match raster to transforms
for (i in 1:Ncovariates) {
   cat(i, covariate.names[[i]],transformation.types[i],"\n")
 }
# Create empty list to file
cov.list <- list()

# Read raster, apply transformation, append to list
for (k in 1:Ncovariates) {
  
  cat("Processing covariate:",covariate.names[[k]],"...\n")
  cov.current <- raster(
    paste0(
      raster_dir,
      covariate.names[[k]]))
  # Step 1: Ensure both rasters are SpatRaster
  cov.current <- rast(cov.current)  # Convert if cov.current is a RasterLayer
  #buffer.image <- rast(buffer.image)  # Convert if buffer.image is a RasterLayer
  
  # Step 2: Make sure both rasters have the same CRS (coordinate reference system)
  if (crs(cov.current) != crs(pop_u5)) {
    cov.current <- project(cov.current, crs(pop_u5))  # Reproject cov.current to match buffer.image's CRS
  }
  
  # Step 3: Crop or Extend
  # If you want to crop cov.current to the extent of buffer.image
  cov.current <- crop(pop_u5, cov.current)  # Crop to the extent of buffer.image
  
  # If you want to extend cov.current to match the extent of buffer.image
  cov.current <- extend(cov.current, ext(pop_u5))  # Extend to the extent of buffer.image
  
  if (!prod(dim(cov.current)==dim(pop_u5))) 
  {warning("Mismatched dimensions!\n")
    cov.current <- resample(cov.current, pop_u5)
    plot(cov.current)}
  cov.current <- values(cov.current)[valid.pix]
  
  if (transformation.types[k]=="Normal") {
    cov.current <- (cov.current-mean(cov.current,na.rm=T))/sd(cov.current,na.rm=T)
  }
  
  if (transformation.types[k]=="Exponential") {
    cov.current <- cov.current + min(
      cov.current[cov.current>0 & cov.current!=Inf & 
                    cov.current!=-Inf & !is.nan(cov.current)], na.rm=T) + 
      abs(min(
        cov.current[cov.current>0 & cov.current!=Inf & cov.current!=-Inf & 
                      !is.nan(cov.current)],na.rm=T))
    cov.current <- qnorm(pexp(cov.current,1/mean(cov.current,na.rm=T)))
  }
  
  nas <- which(is.na(cov.current))
  if (length(nas)>0) {
    validp  <- which(!(is.na(cov.current) | cov.current==-Inf | cov.current==Inf))
    vcoords <- valid.coords[validp,]
    for (i in nas) {
      nearestv <- which.min(
        (vcoords[,1]-valid.coords[i,1])^2+(vcoords[,2]-valid.coords[i,2])^2)
      cov.current[i] <- cov.current[validp[nearestv]]
    }
  }
  
  infs <- which(cov.current==Inf)
  if (length(infs)>0) {
    validp  <- which(!(cov.current==-Inf | cov.current==Inf))
    vcoords <- valid.coords[validp,]
    for (i in infs) {
      nearestv <- which.min(
        (vcoords[,1]-valid.coords[i,1])^2+(vcoords[,2]-valid.coords[i,2])^2)
      cov.current[i] <- cov.current[validp[nearestv]]
    }
  }
  
  neginfs <- which(cov.current==-Inf)
  if (length(neginfs)>0) {
    validp  <- which(!(cov.current==-Inf))
    vcoords <- valid.coords[validp,]
    for (i in neginfs) {
      nearestv <- which.min(
        (vcoords[,1]-valid.coords[i,1])^2+(vcoords[,2]-valid.coords[i,2])^2)
      cov.current[i] <- cov.current[validp[nearestv]]
    }
  }
  # hist(cov.current)
  cov.list[[k]] <- cov.current
}

# Turn list into a matrix
covariates <- do.call(rbind,cov.list)

# Truncate extreme values
covariates[covariates > 5]  <- 5
covariates[covariates < -5] <- -5
```

The Gaussian process implementations we use below assume that the coordinates are on a plane - they don't account for the curvature of the earth (unlike INLA which is able to model GPs on a sphere). We therefore first need to ensure that the raster we use to define this computational setup is appropriately projected. For the block-circulant matrix approach, this is especially important since the computation relies on the compute locations being arranged in a regular gird. We therefore project our matrix to an appropriate projected coordinate reference system:


# Plot the population
```{r pop_plot}
library(terra)
# Ensure pop_raster is a SpatRaster
#Convert SpatRaster to Data Frame
pop_raster_df <- as.data.frame(pop_u5, xy = TRUE)

library(ggplot2)
ggplot()+
  geom_tile(data=pop_raster_df, aes(x = x, y = y,fill = u5_grid3_1km_new))+
  #use na.value to make the lakes appear as lightblue
  scale_fill_viridis_c(option = "magma", na.value = "lightblue", trans = 'log10')+
  theme_minimal()+
  labs(fill = "Population - U5 in Nigeria", fill = "Population Density")

```

Next we prepare our covariates. We will lower the spatial resolution a little to make the example run faster. This is a good idea when initially setting up and debugging a model, but you'll likely want to increase the spatial resolution later. Then we simulate some fake data, using the first 5 covariates:

Read ITN use in LGA level
```{r ITN_use}
library(raster)
library(terra)
itn_stack <- stack(list.files("/mnt/s3/stratification/CountryFolders/Nigeria/Common files/Covariates/1km/Monthly/ITN/20210531_v1/", ".tif", full.names = TRUE))
names(itn_stack) <- month.name[as.numeric(sub("ITN_use_1km_monthly_", "", names(itn_stack)))]
# Select only layers with valid month names (January - December)
months <- c("January", "February", "March", "April", "May", "June", 
            "July", "August", "September", "October", "November", "December")

# Extract matching layers
itn_stack <- itn_stack[[which(names(itn_stack) %in% months)]]
itn_stack <-itn_stack %>%
  resample(pop_u5) %>%  #added resample to align the extents
  crop(pop_u5) %>% 
  mask(pop_u5)

new_crs <-"epsg:26332"  #"epsg:21036"
# project these file
itn_stack <- terra::project(rast(itn_stack), new_crs)

buffer.image <- terra::project(rast(buffer.image), new_crs)

# names(itn_stack) <- month.name[as.numeric(sub("ITN_use_1km_monthly_", "", names(itn_stack)))]
# # 
itn_monthly <- terra::extract(itn_stack, country.shp, 'mean')
 names(itn_monthly) = sub("mean.", "", names(itn_monthly))
# # 
itn_monthly <- as_data_frame(itn_monthly) %>% 
  dplyr::select(January,February,March,April, May, June,July,August,    September,October,November,December)
#itn_monthly <- itn_monthly[,c(1:8, 11:14)]
# 
itn_map <- st_bind_cols(shape, itn_monthly)

# itn_annual_use <- rowMeans(itn_monthly, na.rm = TRUE)
# itn_annual_map <- st_bind_cols(country.shp, itn_annual_use)

ggplot(itn_map)+
  geom_sf( aes(geometry=geometry, fill=itn_annual_use))+
  scale_fill_distiller(palette = "Spectral", na.value = "lightblue")+
  labs(fill = "Annual ITN use ", title = "Nigeria LGAs", subtitle = "MIS 2020, 0-5 years")+
  theme_bw()
```


Plot the clinical case timeseries for all health facilities over each year:

```{r vis_clinical_cases_1}
nga_inc_data <- read.csv("/mnt/efs/stratification/Nigeria/HBHI_2025/03_Additional_files_optional/Imputation/tgo_dhis2_imputed_aggregated.csv")%>%
     arrange(date)

nga_inc_data$date <- as.Date(paste0("01-", nga_inc_data$date), format = "%d-%b %Y")

# Filter out rows with NA values
nga_inc_data_clean <- nga_inc_data[!is.na(nga_inc_data$date) & !is.na(nga_inc_data$positive), ]


 ggplot(
   aes(
     x = month,
     y = positive,
     colour = adm2,
     group = adm2
   ),
   data = nga_inc_data_clean) +
   facet_wrap(~year) +
   geom_line(
     linewidth = 0.3
   ) + 
   theme_minimal() +
   scale_colour_discrete(
     guide = "none"
   ) +
   scale_x_continuous(breaks = 1:12) +
   ggtitle("Clinical case timeseries 0-5 years in Nigeria")
 
 ggsave("~/epiwave.mapping/clinical_cases_timeseries.png")
```

# Make a dynamic plot with ggploty package
```{r vis_clinical_cases_2}
nga_inc_data_tt <- nga_inc_data_clean %>%
  mutate(adm1= case_when(adm1== "federal_capital_territory" ~ "fct",
                         adm1== "abuja_municipal_area_council" ~ "amac",
                         TRUE ~ as.character(adm1) #be sure to include this or it will turn name_1 NA
                         ))

nga_inc_data_tt <- nga_inc_data_clean %>%
  mutate(adm2= case_when(adm2== "abuja_municipal_area_council" ~ "amac",
                         TRUE ~ as.character(adm2) #be sure to include this or it will turn name_1 NA
                         ))

library(plotly)
adm1_plot <- ggplot(
   aes(
     x = month,
     y = positive,
     colour = adm2,
     group = adm1
   ),
   data = nga_inc_data_tt) +
   facet_wrap(~year) +
   geom_line(
     linewidth = 0.3
   ) + 
   theme_minimal() +
   scale_colour_discrete(
     guide = "none"
   ) +
   scale_x_continuous(breaks = 1:12) +
   ggtitle("Clinical case timeseries 0-5 years in Nigeria")
  
p <- ggplotly(adm1_plot)
p
```


Plot the prevalence survey results at the prevalence survey locations:
read prevalence data
```{r}
library(tidyverse)
prev_data <- read_csv("/mnt/s3/stratification/CountryFolders/Nigeria/Projects/1.0_HBHI_risk_mapping/02 Cleaned data/PfPR_NGA_2023.csv") %>% #keep DHS data only...
  janitor::clean_names() %>% 
  mutate(date = make_date(year = year, month = month))

```

```{r vis_prevalence_1}
prev_surveys_mic <- prev_data %>%
  rowwise()%>%
   mutate(n_positive=malaria_test_micro,
          n_sampled = n,
     prevalence = n_positive / n_sampled
   )
# 
 ggplot() +
   geom_sf(data = country.shp) +
   scale_fill_gradient(
     low = grey(0.9),
     high = grey(0.9),
     na.value = "transparent",
     guide = "none") +
   geom_point(
     aes(
       y = latitude,
       x = longitude,
       colour = prevalence
     ),
     data = prev_surveys_mic,
     size = 2,
     shape = 16
   ) +
   scale_colour_gradient(
     labels = scales::label_percent(),
     low = "skyblue",
     high = "darkred"
   ) +
   xlab("") +
   ylab("") +
   theme_minimal() +
   ggtitle("Prevalence surveys with PCR test",
           subtitle="MIS-2020 0-5 years in Nigeria")
```

```{r vis_prevalence_2}
prev_surveys_rdt <- prev_data %>%
  rowwise()%>%
   mutate(n_positive=malaria_test_rapid,
          n_sampled = n,
     prevalence = n_positive / n_sampled
   )
# 
 ggplot() +
   geom_sf(data = country.shp) +
   scale_fill_gradient(
     low = grey(0.9),
     high = grey(0.9),
     na.value = "transparent",
     guide = "none") +
   geom_point(
     aes(
       y = latitude,
       x = longitude,
       colour = prevalence
     ),
     data = prev_surveys_rdt,
     size = 2,
     shape = 16
   ) +
   scale_colour_gradient(
     labels = scales::label_percent(),
     low = "skyblue",
     high = "darkred"
   ) +
   xlab("") +
   ylab("") +
   theme_minimal() +
   ggtitle("Prevalence surveys with RDT test",
           subtitle="MIS-2020 for 0-5 years in Nigeria")
 ggsave("~/epiwave.mapping/Prevalence_rdt.png")
```

Plot the fake delay distribution (probability mass function over days) and prevalence survey detectability functions:

```{r vis_data_4}
#  max_case_delay <- 14
# case_delay_plot <- tibble(
#    days = seq(0, case_delay_max_days)) %>%
#    mutate(pmf = data$surveillance_information$case_delay_distribution_daily_fun
#  (days)
#   )
# 
# ggplot(
#   aes(x = days,
#       y = pmf),
#   data = case_delay_plot) +
#   geom_step() +
#   theme_minimal() +
#   ggtitle("Case reporting delay distribution",
#           "Probability of reporting X days after infection")
```

```{r vis_data_5}
# detectability_plot <- tibble(
#   days = seq(0, data$surveillance_information$prev_detectability_max_days)) %>%
#   mutate(
#     detectability = data$surveillance_information$prev_detectability_daily_fun
# (days)
#   )
# 
# ggplot(
#   aes(x = days,
#       y = detectability),
#   data = detectability_plot) +
#   geom_step() +
#   theme_minimal() +
#   ggtitle("Prevalence survey detectability function",
#           "Probability of registering a positive if tested X days after infection")
```

Plot the assumed infection prevalence - clinical case incidence relationship:

```{r vis_data_6}
# prev_inc_plot <- tibble(prevalence = seq(0, 0.5, length.out = 100)) %>%
#   mutate(
#     clinical_incidence = data$surveillance_information$prev_inc_function
# (prevalence)
#   )
# 
# ggplot(
#   aes(x = prevalence,
#       y = 365 * clinical_incidence),
#   data = prev_inc_plot) +
#   geom_line() +
#   theme_minimal() +
#   xlab("Prevalence") +
#   ylab("Clinical incidence (annual)") +
#   ggtitle("Prevalence - clinical incidence function")
```

### Model specification using greta

To specify the model in greta, first we define the model hyperparameters with priors, then we create any random effects (here the spatiotemporal random effects `epsilon`) and other model objects that depend on them. Then we can define the observation model - linking our generative model to data. 

First we will define the hyperparameters and spatio-temporal process structure and for the spatio-temporal random effects epsilon. We provide two options: the block-circulant embedding approach, and the sparse GP approach, and we demonstrate how to set up objects for both.

We can use the `epiwave.mapping::deifne_bcb_setup()` function to define the objects required for the block-circulant basis approach to simulating IID Guassian processes over space, and converting them to spatio-temporal Gaussian processes: 

```{r define_bcb}
library(epiwave.mapping)
bcb_setup <- define_bcb_setup(grid_raster)
```

For the sparse GP approach, we instead need to construct a set of spatial 'inducing points' or 'knots' across the region of interest. The number of inducing points controls the accuracy of the approximation - with one inducing point per pixel, this is no longer an approximation. An insufficient number of inducing points will bias inference towards smoother Gaussian processes, though trial and error is required to ensure the inducing point scheme is sufficient to accurately capture the underlying process. For now, we will use a small number of points. We generate points evenly across space in a somewhat ad-hoc way using kmeans clustering. Note that improved efficiency of the approximation could be achieved by weighting inducing points towards more populous areas.

```{r define_inducing}
n_inducing <- 30
pts_sim <- terra::spatSample(grid_raster,
                             n_inducing * 1e3,
                             replace = TRUE,
                             xy = TRUE,
                             values = FALSE,
                             na.rm = TRUE)
                  
inducing <- kmeans(pts_sim, n_inducing)$centers
```

We can plot these to see where they fall and check their coverage:

```{r plot_inducing}
# plot the health facility locations
library(tidyterra)
library(ggplot2)
ggplot() +
  geom_spatraster(data = buffer.image) +
  scale_fill_gradient(
    low = grey(0.9),
    high = grey(0.6),
    transform = "log1p",
    na.value = "transparent",
    guide = "none") +
  geom_point(
    aes(
      y = y,
      x = x
    ),
    data = inducing,
    shape = 16
  ) +
  xlab("") +
  ylab("") +
  theme_minimal() +
  ggtitle("Inducing point locations")
```

Now we can define a space-time Gaussian process over `epsilon`, by creating an isotropic spatial kernel object using the greta.gp R package, a parameter for the AR(1) temporal kernel parameter, and a marginal variance parameter, and passing these to the `epiwave.mapping` functions `sim_gp_bcb()` (for the block-circulant embedding) or `sim_gp_sparse()`.

We first need to define priors for the three hyperparameters of the spatio-temporal process. We define a penalised complexity prior (half-normal) for the marginal variance parameter `sigma`, which shrinks the degree of spatio-temporal correlation towards 0. Since we believe negative residual correlation between time periods (oscillation between negative and positive on a monthly timesteps) to be incredibly unlikely, we define a standard uniform prior on the temporal correlation parameter `theta` so that is positive, but otherwise not informative (the spatial pattern could be either constant, or constantly changing). For the spatial range parameter `phi`, we need to take more care to consider reasonable values in terms of the units of the map (metres). We set up an informative prior to control both the lower and the upper tail of the `phi` - making very short and very long lengthscales (very short-range and long-range spatial correlations) *a priori* unlikely. This helps to resolve the various types of non-identifiability inherent in model-based geostatistics: between the lengthscale and marginal variance parameters, between shorter lengthscales and the spatial covariate parameters `beta` (spatial confounding), and between longer lengthscales and the intercept parameter `alpha`. We do so, by having the user specify upper and lower bounds on the degree of spatial correlation that would be plausible, in terms of the spatial extent of the country under consideration.

```{r spatial_priors}
library(greta)
library(greta.gp)

# Variance of space-time process shrunk towards 0
sigma <- normal(0, 1, truncation = c(0, Inf))
# Temporal correlation forced to be positive, but equally as likely to be strong
# as weak
theta <- uniform(0, 1)

# get the maximum spatial extent of each dimension of the raster, in metres
grid_extent <- as.vector(ext(grid_raster))
x_range <- grid_extent["xmax"] - grid_extent["xmin"]
y_range <- grid_extent["ymax"] - grid_extent["ymin"]
max_distance <- max(x_range, y_range)

# First, find a lower bound for the lengthscale parameter phi, a reasonable
# maximum level of complexity, such that spatial correlation would be 0.1 at
# some fraction of the maximum distance, under a Matern 5/2 covariance
# structure:
phi_lower <- matern_lengthscale_from_correl(distance = max_distance / 5,
                                            correlation = 0.1,
                                            nu = 5/2)

# Do the same for an upper bound. A reasonable spatial correlation for
# some longer distance.
phi_upper <- matern_lengthscale_from_correl(distance = max_distance / 2,
                                            correlation = 0.1,
                                            nu = 5/2)

# Now define a gamma prior on the lengthscale parameter phi, such that the mass
# is low on lengthscales below the lower bound (very complex spatial effects)
# and low on lengthscales above the upper bound (completely flat spatial
# effects). Specifically, we find gamma parameters such that there is only a 10%
# chance of *more complex* spatial correlations than the one above. Note that
# whilst an inverse gamma has the inherent property that it has no mass at 0,
# that is taken care of here by our lower bound, and the gamma is easier to
# solve for these constraints.
gamma_prior_params <- find_gamma_parameters(upper_value = phi_upper,
                                            upper_prob = 0.1,
                                            lower_value = phi_lower,
                                            lower_prob = 0.1)

# # we can double check this converged OK:
# round(pgamma(phi_lower,
#              shape = gamma_prior_params$shape,
#              rate = gamma_prior_params$rate), 2)
# round(1 - pgamma(phi_upper,
#                  shape = gamma_prior_params$shape,
#                  rate = gamma_prior_params$rate), 2)

# define the prior over the lengthscale parameter phi
phi <- gamma(shape = gamma_prior_params$shape,
             rate = gamma_prior_params$rate)

```

```{r prior_hist}
prior_hist <- function(x, title = "Histogram of Prior"){
  sims <- calculate(x, nsim = 1000)
  hist(sims$x,main=title, breaks = 100)
}
```

```{r Plot_priors_histogram}
library(bayesplot)

par(mfrow = c(2, 1))  # Create a 2x1 layout (2 plots)
prior_hist(sigma, "Histogram of sigma")

prior_hist(theta, "Histogram of theta")
```

To use the block-circulant approach we need to define a spatial kernel in greta.gp, with only a single lengthscale parameter, and pass this to `epiwave.mapping::sim_gp_bcb()` along with our setup for the BCB approach. I've commented this out (as well as not evaluating the block), since running this block in addition. to the sparse approach we use below will make the model run very slowly.

```{r define_epsilon_bcb, eval = FALSE}
# # We define an isotropic kernel in greta.gp. We fix the variance of this
# # kernel to 1, since we define that parameter at the level of the combined
# # space-time kernel
# k_space <- mat52(lengthscales = phi,
#                  variance = 1)
# 
# # Now we can pass this kernel and the other hyperparameters to the function to
# # create space-time GPs over all cells
# epsilon <- epiwave.mapping::sim_gp_bcb(
#   bcb_setup = bcb_setup,
#   n_times = n_years * 12,
#   space_kernel = k_space,
#   time_correlation = theta,
#   sigma = sigma
# )
```

Alternatively, we can construct the space-time process using a sparse GP with `epiwave.mapping::sim_gp_sparse()`. In this case (*for reasons*), to specify an isotropic spatial kernel we need to pass the lengthscale parameter to the `greta.gp` kernel function twice, once for each dimension:
```{r define_epsilon_sparse}
# note we pass the lengthscale argument twice
k_space <- mat52(lengthscales = c(phi, phi),
                 variance = 1)

# set time periods
start_year <- 2021
n_years <- 3

epsilon <- epiwave.mapping::sim_gp_sparse(
  inducing_points = inducing,
  grid_raster = grid_raster,
  n_times = n_years * 12,
  space_kernel = k_space,
  time_correlation = theta,
  sigma = sigma
)

```

We can visually check that our prior is giving us reasonable values of epsilon by doing some prior simulations and then plotting them:
```{r gp_prior_check, fig.width = 10, fig.height=9}
n_sims <- 3
times_plot <- 1:5
prior_sims <- calculate(epsilon,
                        phi,
                        theta,
                        sigma,
                        nsim = n_sims)

# for each simulation, create an empty multilayer raster, fill it in with the
# simulations for each of the first three years, and plot it
plot_list <- list() 
for (sim in seq_len(n_sims)) {
  
  plot_raster <- do.call(c,
                         replicate(length(times_plot),
                                   grid_raster,
                                   simplify = FALSE))
  names(plot_raster) <- paste("timestep", times_plot)
  
  for (i in seq_along(times_plot)) {
    plot_raster[[i]][cells(grid_raster)] <- prior_sims$epsilon[sim, , i]
  }

  plot_list[[sim]] <- ggplot() +
    geom_spatraster(data = plot_raster) +
    scale_fill_distiller(
      palette = "PiYG",
      na.value = "transparent",
      guide = "none") +
    facet_wrap(~lyr, nrow = 1) +
    theme_minimal() +
    xlab("") +
    ylab("") +
    ggtitle(label = paste("Simulation", sim),
            subtitle = sprintf(
              "phi (space) = %s, theta (time) = %s, sigma = %s",
              round(prior_sims$phi[sim, 1, 1]),
              round(prior_sims$theta[sim, 1, 1], 2),
              round(prior_sims$sigma[sim, 1, 1], 2)
            ))
}

library(patchwork)
combined_plot <- plot_list[[1]]
for (sim in 2:n_sims) {
  combined_plot <- combined_plot / plot_list[[sim]]
}
combined_plot
```
You can run that block multiple times to see different realisations, and check they conform to you prior expectation about the spatio-temporal random process.

Now we have the object for the spatio-temporal random effects, we define our remaining hyperparameters for the covariate effects, and reporting fraction.

Here we use an arbitrary set of priors, that happen to be those used to simulate the 'true' parameters in `epiwave.mapping::sim_data()`. In a practical application, the user should give these a lot more thought. 
```{r define_priors}
# intercept and slope for fixed-effects terms on daily infection incidence
alpha <- normal(log(1e-4), 1)
beta <- normal(0, 1, dim = terra::nlyr(covariates_raster))
# fraction of all clinical cases reported in case counts
r <- beta(10, 5)

prior_hist(alpha, "histogram of alpha")
prior_hist(beta, "histogram of beta")
prior_hist(r, "histogram of r")
```

Then we can define the deterministic mapping from these, through infection incidence, to our observed data. First we compute the daily incidence of new infections, per cell, per timestep.
```{r define_infection_incidence}

# get an index to all pixel values
all_cells <- terra::cells(grid_raster)

# extract the covariate design matrix (scale)
X <- terra::extract(covariates_raster, all_cells)

    # Check for NA values
any(is.na(X))
# Check for NAs for a single column
sum(is.na(X$LST.Day))

# find an index to all the cells that have data
#not_na_idx <- which(!is.nan(X$LST.Day))

# Replace NA values with 0 (or another meaningful value)
X[is.na(X)] <- 0

# matrix-multiply with beta, and add alpha, to get the spatial fixed effects
fixef <- alpha + X %*% beta # we have to add the ITN use: -b_ITN*Y_ITN

# add on epsilon, and convert to infection incidence
eta <- sweep(epsilon, 1, fixef, FUN = "+")
infection_incidence <- ilogit(eta)

# multiply by population and timeperiod to get the count of new infections per
# time period
timeperiod_days <- 365/12
pop_all_cells <- terra::extract(buffer.image, all_cells)[[1]]

daily_new_infections <- sweep(infection_incidence, 1, pop_all_cells, FUN = "*")
timeperiod_new_infections <- daily_new_infections * timeperiod_days
```   
Note that we can similarly simulate other objects like `timeperiod_new_infections` from the model priors and plot them as rasters to check the model and prior specification seems reasonable.

Next we will compute the expected prevalence at our prevalence survey locations. We need to find the index to the rows in `timeperiod_new_infections` where the prevalence surveys were performed, and then apply the appropriate convolution on these to compute the expeccted prevalences at those times and places.

```{r}
library(sf)
library(raster)

# Example: Your prevalence dataframe (make sure it has "longitude" and "latitude" columns)

# Convert the dataframe into an `sf` object
prevalence_sf <- st_as_sf(prev_surveys_rdt,
                          coords = c("longitude", "latitude"),
                          crs = 4326) %>%
  sf::st_transform(crs = st_crs(covariates_raster[[1]]))


# Print to check if it worked
print(prevalence_sf)

# Check the CRS
st_crs(prevalence_sf) == st_crs(grid_raster)  # Should return TRUE

```


```{r define_prevalence}
# get the prevalence survey coordinates (already projected int he same CRS as the rasters)
prev_coords <- prevalence_sf %>%
  st_coordinates()

# find an index to the cells in the full grid
prev_cells <- terra::cellFromXY(grid_raster, prev_coords)
# get the index to rows in our compute objects
prev_index <- match(prev_cells, all_cells)
prev_index <- prev_index[!is.na(prev_index)]
# pull out the number of infections per day in these locations
daily_new_infections_prev_locs <- daily_new_infections[prev_index, ]

#*********************************************
#Add by OD
# for each day post-infection this is the (entirely fake) probability of testing
  # positive on our hypothetical test, if tested that day. Hard-coded to be 0 for
  # negative times or beyond 30 days
  max_infection_detectability <- 30
  q_daily <- function(days) {
    up <- plogis((days * 2) - 5)
    down <- 1 - plogis(days / 2 - 10)
    q <- up  * down
    q[days < 0] <- 0
    q[days > 30] <- 0
    q
  }
#****************************************
# now to do the convolution, we first need to translate the detectability
# function from a daily timestep to our monthly timestep
#q_daily <- data$surveillance_information$prev_detectability_daily_fun
q_max_days <- 30 #data$surveillance_information$prev_detectability_max_days
timeperiod_days
q_timeperiod <- transform_convolution_kernel(kernel_daily = q_daily,
                                  max_diff_days = q_max_days,
                                  timeperiod_days = timeperiod_days)

# now we can do the convolution to get the number of detectable infections, if
# measured each day
detectable_infections_daily_prev_locs <- convolve_matrix(
  matrix = daily_new_infections_prev_locs,
  kernel = q_timeperiod)

# compute prevalence at these locations over time by dividing by population of
# the locations where prevalence surveys were done
population_prev_locs <- pop_all_cells[prev_index]
prevalence_prev_locs <- sweep(detectable_infections_daily_prev_locs,
      1, population_prev_locs, FUN = "/")
```

Now we model the expected clinical incidence at each pixel, accounting for the prevalence-incidence relationship, delays between infection and reporting, imperfect reporting rates, and aggregating the counts over health facility catchments.

```{r define_clinical_incidence}
# first, we define gamma: the ratio of clinical cases to infections

# We estimate q_star, the integral of the daily detectability function
q_star <- sum(q_daily(seq(0, q_max_days)))

# As per equations above, we can get the instantaneous prevalence using this,
# and the provided prevalence-incidence relationship to map to the incidence
# rate of all clinical cases (reported or not)
# create a function g for the prevalence-incidence relationship
  g <- function(prevalence) {
    (1/365) * (1 - exp(-prevalence * 10))
  }
#g <- data$surveillance_information$prev_inc_function
clinical_case_incidence <- g(q_star * infection_incidence)
# multiplying by the reporting rate r gives us the reported clinical incidence rate
reported_clinical_case_incidence <- r * clinical_case_incidence

# to convert this to counts of reported clinical cases per pixel (before
# temporal convolution), we compute and use gamma like in the above equations,
# like this:
gamma <- reported_clinical_case_incidence / infection_incidence
reported_clinical_cases <- gamma * timeperiod_new_infections

# note we could also multiply reported_clinical_case_incidence by the
# populations and the monthly timeperiod

# next, we convolve these through time to account for delays between infection
# and reporting. This follows the same process as for prevalence:

#pi_daily <- data$surveillance_information$case_delay_distribution_daily_fun
# now do the same for the pixel map of clinical case counts
pi_daily <- function(day_difference) {
    upper_limit <- max_case_delay
    # use a truncated lognormal (truncated to 0-upper_limit)
    mu <- 1.2
    sigma <- 0.5
    day_difference_old <- day_difference
    day_difference <- round(day_difference)
    if (!max(abs(day_difference - day_difference_old) < 1e-4)) {
      warning("day_difference was non-integer, using rounded values")
    }

    upper <- plnorm(day_difference + 1, mu, sigma)
    lower <- plnorm(day_difference, mu, sigma)
    dens <- upper - lower

    # correct the densities
    norm <- plnorm(upper_limit, mu, sigma)
    res <- dens / norm
    res[day_difference > upper_limit] <- 0
    res[day_difference < 0] <- 0
    res
  }

pi_max_days <- max_case_delay <- 14 #data$surveillance_information$case_delay_max_days
pi_timeperiod <- transform_convolution_kernel(kernel_daily = pi_daily,
                                              max_diff_days = pi_max_days,
                                              timeperiod_days = timeperiod_days)

# now we can convolve to compute prevalence at all pixels over time
reported_clinical_cases_shifted <- convolve_matrix(reported_clinical_cases,
                                        kernel = pi_timeperiod)
```


This gives us the expected number of clinical cases reported in each
timeperiod, in each pixel. Next we need to aggregate them at health facility
level to link them to the available data.

```{r health_facility_aggregation}
# # First, we'll need to reproject the catchment model, and then extract all
# # catchment information into a matrix
# catchment_rast_raw <- data$surveillance_information$health_facilities$catchments_rast
# catchment_rast <- terra::project(catchment_rast_raw, new_crs) 
# catchment_weights <- terra::extract(catchment_rast, all_cells)
# 
# # after reprojecting, we will need to ensure that the weights sum to 1 across
# # all rows(pixels), so that the weights are probabilities of attending each
# # health facility. If they don't sum to 1, we'll be losing/adding cases in an
# # unpredictable way
# catchment_weight_sums <- rowSums(catchment_weights)
# catchment_weights <- sweep(catchment_weights, 1, catchment_weight_sums, FUN = "/")
# 
# # now we just need a matrix multiplication to assign the reported cases in each
# # pixel and time to health facility level, resulting in a matrix with rows
# # giving health facilities and columns giving time periods
# reported_clinical_cases_hf <- t(catchment_weights) %*% reported_clinical_cases_shifted
```

# As we are working on LGAs, the we do not take into account the catchments
# Then, we define here how to aggregate at that level
```{r lga_aggregation}
# First, we'll need to reproject the catchment model, and then extract all
# catchment information into a matrix
#catchment_rast_raw <- data$surveillance_information$health_facilities$catchments_rast
#catchment_rast <- terra::project(catchment_rast_raw, new_crs) 
catchment_weights <- terra::extract(reported_clinical_cases_shifted, all_cells,'mean')

# after reprojecting, we will need to ensure that the weights sum to 1 across
# all rows(pixels), so that the weights are probabilities of attending each
# health facility. If they don't sum to 1, we'll be losing/adding cases in an
# unpredictable way
catchment_weight_sums <- rowSums(catchment_weights)
catchment_weights <- sweep(catchment_weights, 1, catchment_weight_sums, FUN = "/")

# now we just need a matrix multiplication to assign the reported cases in each
# pixel and time to health facility level, resulting in a matrix with rows
# giving health facilities and columns giving time periods
reported_clinical_cases_hf <- t(catchment_weights) %*% reported_clinical_cases_shifted
```


Now we define likelihoods for the two types of observation data. First we define the likelihood for prevalence survey data by extracting the prevalence values from the modelled timeseries, at the times and locations where the surveys were performed.
```{r prevalence_likelihood}
# Now we find the timeperiod in which each survey was done, and extract the
# corresponding prevalence estimates
n_surveys <- nrow(data$epi_data$prevalence_surveys)
idx <- cbind(
  seq_len(n_surveys), # rows
  data$epi_data$prevalence_surveys$time # columns
)

prevalence_estimates <- prevalence_prev_locs[idx]

# now we can define a binomial likelihood over the data for this part of the
# model
positive <- data$epi_data$prevalence_surveys$n_positive
tested <- data$epi_data$prevalence_surveys$n_sampled
distribution(positive) <- binomial(tested, prevalence_estimates)
```

Now we add the likelihood for the clinical cases reported at health facility level, accounting for missingness of data in these timeseries.
```{r clinical_incidence_likelihood}
# find the non-missing data
valid_idx <- !is.na(data$epi_data$clinical_cases$cases)

# pull out these places, times, and cases
cases_valid <- data$epi_data$clinical_cases[valid_idx, ]
cases_observed <- cases_valid$cases

# extract the corresponding estimates of cases
cases_extract_index <- cbind(cases_valid$health_facility, cases_valid$time)
cases_expected <- reported_clinical_cases_hf[cases_extract_index]

# define the poisson observation model
distribution(cases_observed) <- poisson(cases_expected)
```

### Model fitting and prediction

Now we have defined all of the components of the model, we can run MCMC to estimate parameters.

In general, it would be a good idea to use calculate to simulate some of these quantities (predicted prevalences etc) from the priors, to make sure they are somewhat reasonable, and there are no numerical issues or mistakes in the model code. I'll skip that for now in this example, since the model has already been checked.

```{r run_mcmc}
# list only the quantities/parameters we want to trace the posteriors for in the
# first instance. We can always recompute other quantities later, so keep this
# to the hyperparameters.
m <- model(
  alpha, beta,
  r,
  phi, theta, sigma
)

# the model has trouble initialising when the infection incidence is high
# (because the number of people detectable in the prevalence surveys can exceed
# the population), so initialise it to be low

# # later: work out how to cap it without ruining the gradients
n_chains <- 4
inits <- replicate(n_chains,
                   initials(
                     alpha = runif(1, -10, -9)
                   ),
                   simplify = FALSE)

draws <- mcmc(m,
              chains = n_chains,
              initial_values = inits)

r_hats <- coda::gelman.diag(draws, autoburnin = FALSE, multivariate = FALSE)
```

Now we have sampled the model, we can plot posterior maps of quantities we are interested in. Here we will the infection incidence rate (annual, per person) and the numbers of clinical cases expected per pixel, per timeperiod. To get posterior samples of quantities of interest for greta arrays we pass the posterior samples to the `values` argument of tha `greta::calculate()` function. Then we summarise these to get the posterior means before putting them in the rasters to visualise them.
```{r plot_posteriors, fig.width = 10, fig.height=6}

# first, we calculate the number of clinical cases (reported or otherwise) at the time of reporting, per timeperiod, per pixel
# clinical_cases_pixel <- reported_clinical_cases_shifted / r
clinical_cases_pixel <- reported_clinical_cases / r

posterior_sims <- calculate(infection_incidence,
                            clinical_cases_pixel,
                            values = draws,
                            nsim = 1000)

# in these, the first dimension is the posterior sample, the second is the pixel location, and the third is the timeperiod. Sowe calculate the mean across the first dimension
infection_incidence_post_mean <- apply(posterior_sims$infection_incidence, 2:3, mean)
clinical_cases_pixel_post_mean <- apply(posterior_sims$clinical_cases_pixel, 2:3, mean)

# now we can put these in rasters
n_times <- ncol(infection_incidence)
plot_raster <- do.call(c,
                       replicate(n_times,
                                 grid_raster,
                                 simplify = FALSE))
names(plot_raster) <- paste("timestep", seq_len(n_times))

infection_incidence_post_raster <- plot_raster
clinical_cases_pixel_post_raster <- plot_raster
for (i in seq_len(n_times)) {
  infection_incidence_post_raster[[i]][all_cells] <- infection_incidence_post_mean[ , i]
  clinical_cases_pixel_post_raster[[i]][all_cells] <- clinical_cases_pixel_post_mean[ , i]
}

# plot for 5 timeperiods
times_plot <- round(seq(1, n_times, length.out = 5))

infection_incidence_plot <- ggplot() +
    geom_spatraster(data = 365 * infection_incidence_post_raster[[times_plot]]) +
    scale_fill_distiller(
      name = "Incidence",
      palette = "YlGnBu",
      direction = 1,
      na.value = "transparent") +
    facet_wrap(~lyr, nrow = 1) +
    theme_minimal() +
    xlab("") +
    ylab("") +
    ggtitle("Infection incidence",
            "per person per year")

clinical_cases_pixel_plot <- ggplot() +
    geom_spatraster(data = clinical_cases_pixel_post_raster[[times_plot]]) +
    scale_fill_distiller(
      name = "Cases",
      palette = "YlOrRd",
      direction = 1,
      na.value = "transparent") +
    facet_wrap(~lyr, nrow = 1) +
    theme_minimal() +
    xlab("") +
    ylab("") +
    ggtitle("Clinical cases",
            "number per pixel per timeperiod")

clinical_cases_pixel_plot / infection_incidence_plot
```

### Comparison with simulated truth

We can compare these estimates with the 'true' simulated surfaces from the `data` object.
```{r plot_truth, fig.width = 10, fig.height=6}
infection_incidence_truth_raster <- data$truth$rasters$infection_incidence_rast

clinical_cases_truth_raster <- data$truth$rasters$clinical_cases_rast


# plot for 5 timeperiods
times_plot <- round(seq(1, n_times, length.out = 5))

infection_incidence_truth_plot <- ggplot() +
    geom_spatraster(data = 365 * infection_incidence_truth_raster[[times_plot]]) +
    scale_fill_distiller(
      name = "Incidence",
      palette = "YlGnBu",
      direction = 1,
      na.value = "transparent") +
    facet_wrap(~lyr, nrow = 1) +
    theme_minimal() +
    xlab("") +
    ylab("") +
    ggtitle("Infection incidence",
            "per person per year")

clinical_cases_truth_plot <- ggplot() +
    geom_spatraster(data = clinical_cases_truth_raster[[times_plot]]) +
    scale_fill_distiller(
      name = "Cases",
      palette = "YlOrRd",
      direction = 1,
      na.value = "transparent") +
    facet_wrap(~lyr, nrow = 1) +
    theme_minimal() +
    xlab("") +
    ylab("") +
    ggtitle("Clinical cases",
            "number per pixel per timeperiod")

clinical_cases_truth_plot / infection_incidence_truth_plot
```

We can also compare the posterior estimates of the hyperparameters against their true values in the simulation
```{r compare_params}

# get true values
param_names <- c("alpha", "beta", "r", "phi", "theta", "sigma")
param_truth <- unlist(data$truth$parameters[param_names])

# compute summaries of parameters in the same order
param_post <- calculate(
  alpha,
  beta,
  r,
  phi,
  theta,
  sigma,
  values = draws,
  nsim = 1e3
)

param_prior <- calculate(
  alpha,
  beta,
  r,
  phi,
  theta,
  sigma,
  nsim = 1e3
)

post_mean <- unlist(lapply(param_post, colMeans))
post_lower <- unlist(lapply(param_post, function(x) apply(x, 2:3, quantile, 0.025)))
post_upper <- unlist(lapply(param_post, function(x) apply(x, 2:3, quantile, 0.975)))

prior_mean <- unlist(lapply(param_prior, colMeans))
prior_lower <- unlist(lapply(param_prior, function(x) apply(x, 2:3, quantile, 0.025)))
prior_upper <- unlist(lapply(param_prior, function(x) apply(x, 2:3, quantile, 0.975)))

tibble(
  parameter = names(param_truth),
  truth = round(param_truth, 2),
  posterior_mean = round(post_mean, 2),
  posterior_ci = paste(
    round(post_lower, 1),
    round(post_upper, 1),
    sep = " to "
  ),
  prior_ci = paste(
    round(prior_lower, 1),
    round(prior_upper, 1),
    sep = " to "
  ),
)

```
