---
output:
  github_document:
    pandoc_args: --webtex
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# epiwave.mapping

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Prototype R package and example code for computationall-yefficient, fully-Bayesian semi-mechanistic spatio-temporal mapping of disease infection incidence simultaneously from spatially-aggregated (e.g. at health facility level) clinical case count timeseries, and targeted infection prevalence surveys at specific point locations. This model and software is based on ideas in the epiwave R package and uses some functions implemented there, but it currently relies on the user to develop their own model using the greta R package, rather than providing wrapper functions to set up the model for the user.

## Installation

You can install the package from github, using the `remotes` package:

```{r install, eval = FALSE}
remotes::install_github("idem-lab/epiwave.mapping",
                        dependencies = TRUE)
```

## Model

Our main aim is to model, and generate spatio-temporal predictive maps of, variation in *infection incidence*: the average number of new infections per time period, per member of the population. We model the logarithm of infection incidence with a space-time Gaussian process. Since infection incidence is rarely perfectly observed, we infer it from two independent and complementary datastreams:

- **Clinical case timeseries**: Counts of the numbers of cases reported over time, but potentially spatially aggregated across e.g. the catchment areas of health facilities, or the entire region, rather than at spatially precise coordinates

- **Infection prevalence survey points**: Random samples of individuals at a specific point location, with the number tested and number positive recorded. Typically these data are available only very infrequently.

Since neither datastream provides a direct estimate of infection incidence, we model the generative observation process yielding each datastream, using informative priors for key parameters wherever possible. A number of options are available for computationally-efficient space-time Gaussian process (GP) modelling for applications of this type. Below we implement an approach well-suited to our health facility catchment observation model and fully Bayesian inference using Hamiltonian Monte Carlo: a separable space-time GP employing a block-circulant embedding for the spatial process. This is similar to the approach used in the `lgcp` R package, and detailed in Diggle et al. (2013). We note that the clinical incidence observation model we employ is a particular case of a Log-Gaussian Cox process. Our contribution is accounting for delays in reporting, and linking model this to prevalence survey data, in a practical and easily-extensible framework.

### Clinical incidence model 

We model the observed number of clinical cases $C_{h,t}$ of our disease of interest in health facility $h$ during discrete time period $t$ as a Poisson random variable:

$$
C_{h,t} \sim \text{Poisson}(\hat{C}_{h,t})
$$
The expectation of this Poisson random variable (the modelled/expected number of clinical cases) is given by a weighted sum of (unobserved but modelled) expected pixel-level clinical case counts $\hat{C}_{l,t}$ at each of $L$ pixel locations $l$:

$$
\hat{C}_{h,t} = \sum_{l=1}^{L}{\hat{C}_{l,t}w_{l,h}}
$$

where weights $w_{l,h}$ give the 'membership' of the population in each pixel location to each health facility, where each case at a given location $l$ has probability $w_{l,h}$ of reporting at health facility $h$, and therefore $\sum_{l=1}^L w_{l,h} = 1$. In practice this could be either a proportional (fractions of the population attend different health facilities) or a discrete (the population of each pixel location attends only one nearby health facility) mapping.

At each pixel location, we model the unobserved clinical case count $\hat{C}_{l,t}$ from the modelled number of new infections $\hat{I}_{l,t'}$ at the same location, during previous time periods $t'$, from the fraction of infections $\gamma_{l,t'}$ at that location and previous time period that would result in a recorded clinical case, and a probability distribution $\pi(t-t')$ over delays $t-t'$ from infection to diagnosis and reporting:

$$
\hat{C}_{l,t} = \sum_{t-t' = 0}^{\tau_\pi}{\hat{I}_{l,t'} \, \gamma_{l,t'} \, \pi(t-t')}
$$
The distribution $\pi(\Delta_t)$ gives the discrete and finite (support on $\Delta_t \in (0, \tau_\pi)$) probability distribution over delays from infection to reporting, indexed on the time-periods considered in the model. This temporal reweighting to account for a distribution over possible delays can be considered as a 'discrete-time convolution' with $\pi(\Delta_t)$ the 'kernel'. Below we discuss efficient methods for computing these convolutions in greta.

### Infection prevalence model 

We model the observed number of individuals who test positive for infection $N^+_{l,t}$ in an infection prevalence survey at location $l$ at time $t$ as a binomial sample, given the number of individuals tested $N_{l,t}$, and the modelled prevalence of infections $\hat{p}_{l,t}$ across the population at that time/place:

$$
N^+_{l,t} \sim \text{Binomial}(N_{l,t}, \hat{p}_{l,t})
$$
Similarly to clinical incidence, we model the infection prevalence at a given location and time as a discrete-time convolution over previous infection counts, divided by the total population in the pixel $M_l$, and the duration of the time-period in days $d$:

$$
\hat{p}_{l,t} = \frac{1}{M_l}\sum_{t-t' = 0}^{\tau_q}{q(t-t') \, \hat{I}_{l,t'} \, d^{-1}}
$$

In this case, the kernel $q(t-t')$ is not a probability distribution, but the kernel of a convolution operation that maps the daily average number of new infections in previous time periods $t'$ to the number of individuals who would test positive (to the diagnostic used) on a randomly-selected day in the timeperiod `t`. This kernel, which is specific to the tim eperiod chosen for modelling the temporal process, can be computed from a related function: $q_{\text{daily}}(\Delta_t)$ which gives the probability that an individual will test positive $\Delta_t$ days after infection. The integral of this function is the average number of days an infected person would test positive for. The function $q_{\text{daily}}()$ can be estimated from empirical data on how the test sensitivity and duration of infection vary over time since infection. This package provides functionality: `transform_convolution_kernel()` to construct timeperiod-specific kernels such as $q()$ from daily kernels like $q_{\text{daily}}()$ for an arbitrary modelling timeperiod.

By convolving of the number of new infections *per day* $\hat{I}_{l,t'} \, d^{-1}$ in all previous time periods with the fraction of those that would test positive in a survey in time period $t$, gives an estimate of the *number* of positive-testing people in the population at location $l$, on any given day within the time period $t$. Dividing this by the population of that location, $M_l$, therefore yields an estimate of the population proportion testing positive on any given day; the parameter of the binomial distribution that captures the prevalence survey data.

Note that our definition of $\hat{p}_{l,t}$ is the population prevalence of infections *that would test positive using that diagnostic method* rather than the true fraction infected/infectious at any one time. We also assume here that tests have perfect specificity, though the model can easily be adapted to situations where that is not the case.

### Infection incidence model 

The expected number of new infections $\hat{I}_{l,t'}$ in location $l$ during time period $t$ is modelled as the product of the population $M_l$ at that location, the length of the timeperiod in days $d$, and the daily infection incidence $f_{l,t}$ at that location and time:

$$
\hat{I}_{l,t'} = d \, M_l \, \hat{f}_{l,t}
$$

Whilst we mechanistically model the observation processes yielding our data types, we employ a geostatistical approach to modelling spatio-temporal variation in infection incidence, with spatio-temporal covariates $\mathbf{X}_{l,t}$ and a space-time random effect $\epsilon_{l,t}$ with zero-mean Gaussian process prior:

$$
\text{log}(f_{l,t}) = \alpha +\mathbf{X}_{l,t} \beta + \epsilon_{l,t}
$$

$$
\epsilon \sim GP(0, \mathbf{K})
$$

where $\alpha$ is a scalar intercept term, $\beta$ is a vector of regression coefficients against the environmental covariates, and $\mathbf{K}$ is the space-time covariance function of the Gaussian process over random effects $\epsilon_{l,t}$.

There are many choices of space-time covariance structure for $\mathbf{K}$, though we use a separable combination of an isotropic spatial covariance function with a temporal covariance function, to enable the use of a range of computationally efficient simulation and calculation methods:

$$
K_{l,t,l',t'} = \sigma^2 \, K_{\text{space}}(||l-l'||; \phi) \, K_{\text{time}}(|t-t'|; \theta)
$$
where $\sigma^2$ is the marginal variance (amplitude) of the resultant Gaussian process, $K_{\text{space}}(.; \phi)$ is an (unit variance) isotropic spatial kernel on euclidean distances $||l-l'||$ with parameter $\phi > 0$ controlling the range of spatial correlation, and $K_{\text{time}}(|t-t'|; \theta)$ is a a temporal kernel on time differences $|t-t'|$, with parameter $\theta$ controlling the range of temporal correlation. Again for computational reasons, we prefer Markovian kernels for $K_{\text{time}}$.

## Computational approaches

### Inference

We fit the model via fully Bayesian inference, using MCMC (Hamiltonian Monte Carlo) in the greta R package. Whilst a number of alternative inference algorithms have been proposed and widely adopted for Bayesian and non-Bayesian estimation of spatio-temporal Gaussian process models (e.g. INLA), in out case, the aggregation of clinical cases across catchments (a feature required by the model structure) leads to a non-factorising likelihood, which nullifies many of the computational advantages of that method. Further, the mapping from the Gaussian process to the expectations of the two sampling distributions is non-linear (due to sums of exponents in various places), requiring computationally costly linearisation of the non-linearities. By employing a fully-Bayesian MCMC approach, and focussing on computationally-efficient model and implementation choices, rather than approximations, we are able to estimate the model parameters relatively quickly, with full treatment, propagation, and quantification of model uncertainty, and retaining the ability to easily modify and interrogate the model.

### Discrete-time convolution

This discrete convolutions (weighted sums over previous time points) used to compute $\hat{C}_{l,t}$ and $\hat{p}_{l,t}$ can be computationally intensive, depending on the number of time periods modelled. A number of efficient computational approaches exist to overcome these, and the optimal approach to use in greta depends on the size of $\tau^{max}$: the maximum duration of the delay in terms of the number of timeperiods being considered. Where $\tau^{max}$ is relatively large (say, $\tau^{max} > 3$), implementation of the convolution as a matrix multiply is likely to be the most efficient in greta. If $\tau^{max} > 3$ and the number of time periods being modelled is much larger than $\tau^{max}$, implementing this as a sparse matrix multiply (skipping computation on zero elements of the convolution matrix) is likely to be most efficient. Where $\tau^{max}$ is small (say, $\tau^{max} \leq 3$) the convolution can instead be computed with a sum of dense vectorised additions and subtractions. These convolution approaches are implemented in this package, and demonstrated below.

### Gaussian process simulation

Naive implementation of the full Gaussian process (GP) model is typically very computationally intensive, due to the fact that the most expensive step (inverting a dense covariance matrix) scales cubically with the number of evaluation points. In the model we propose (and as in a Log-Gaussian Cox process), we need to evaluate the GP at every pixel location and time period in the study frame, so that we can aggregate the expected clinical case count to compute the likelihood. This results in a computationally impractical algorithm for all but the smallest study frames.

One solution would be to employ an approximation to the full GP, evaluated at only a subset of locations and times in the study frame, and approximating the clinical case count calculation with some smaller finite sum. Candidate GP approximation approaches include the SPDE approximation to a Matern-type spatial kernel, on a computational mesh (as used in INLA, Lindgren *et al.*, 2011), a sparse GP method over a limited set of inducing points (Quinonera-Candela *et al.*, ?; see `greta.gp`), or one of the closely-related penalised spline methods (see `greta.gam`).

An appealing alternative in this case is to compute the full spatial GP for every pixel in a regular grid (the same we use to record spatial covariate values), by exploiting the block-circulant structure in the resulting spatial covariance matrix. This requires using projected coordinates, and expanding the spatial study frame (by a factor of 2 in each dimension) to map the projection onto a torus in such a way that the distances between pairs of pixels are preserved. This approach enables simulation of the GP across all pixels, for a given time period, via the fast fourier transform (FFT) - which scales only linearly with the number of locations considered (? check when have wifi). This enables us to simulate values of the spatial GP across all pixels very cheaply, with no need to approximate the GP. In separable combination with a Markovian temporl kernel, this enables very rapid inference. 

## Example application

We demonstrate the model with application to mapping the (simulated) infection incidence of malaria in Kenya from simulated clinical incidence and infection prevalence data. Using the `sim_data()` function from this package, we use a real grid of environmental covariates, but simulate the locations of health facilities, prevalence survey locations, and all datasets to which we will then fit the model.

First we load the bioclim covariates using the `terra` and `geodata` R package:
```{r bioclim, eval = FALSE}
library(terra)
library(geodata)
# download at a half-degree resolution. See ?geodata_path for how to save the data between sessions
bioclim_kenya <- geodata::worldclim_country(
  country = "KEN",
  var = "bio",
  res = 0.5,
  path = file.path(tempdir(), "bioclim")
)

pop <- geodata::population(
  year = 2020,
  res = 0.5,
  path = file.path(tempdir(), "population")
)

# crop and mask to Kenya
pop_kenya <- mask(pop, bioclim_kenya)

```

```{r bioclim_secret, eval = TRUE, echo = FALSE}
library(terra)
library(geodata)
# I'm on a plane writing this, so mocking up the chunk with a file I know I have
source_bioc <- "../vector_atlas_training_2024/data/rasters/bc_kenya.tif"
bioclim_kenya <- rast(source_bioc)
source_pop <- "../ir_cube/data/clean/pop_cube.tif"

kenya_mask <- bioclim_kenya[[1]]
pop <- rast(source_pop)$pop_2020
pop_kenya <- crop(pop, kenya_mask)
disagg_factor <- 5
pop_kenya <- terra::disagg(pop_kenya, disagg_factor) / (disagg_factor ^ 2)
pop_kenya <- resample(pop_kenya, kenya_mask)
pop_kenya <- mask(pop_kenya, kenya_mask)
```

Now we lower the spatial resolution. a little (to make the example run faster) and simulate some fake data, using the first 5 covariates
```{r sim_data}
bioclim_kenya_lores <- terra::aggregate(bioclim_kenya, 10)
pop_kenya_lores <- terra::aggregate(pop_kenya, 10)

library(epiwave.mapping)
set.seed(1)
start_year <- 2020
n_years <- 5
data <- sim_data(
  covariates_rast = bioclim_kenya_lores[[1:5]],
  population_rast = pop_kenya_lores, 
  years = seq(start_year, start_year + n_years - 1),
  n_health_facilities = 60,
  n_prev_surveys = 30,
  # assume some fraction of case counts are missing
  case_missingness = 0.3)
```

Let's visualise the fake data.

Plot the fake health facility locations over the population raster:

```{r vis_data_1}
library(tidyverse)
library(tidyterra)

# plot the health facility locations
ggplot() +
  geom_spatraster(data = pop_kenya_lores) +
  scale_fill_gradient(
    low = grey(0.9),
    high = grey(0.6),
    transform = "log1p",
    na.value = "transparent",
    guide = "none") +
  geom_point(
    aes(
      y = y,
      x = x
    ),
    data = data$surveillance_information$health_facilities$health_facility_loc,
    shape = 16
  ) +
  xlab("") +
  ylab("") +
  theme_minimal()
```
Now let's plot the clinical case timeseries

```{r vis_data_2}

# add months and years onto the clinical cases, for plotting
cases <- data$epi_data$clinical_cases %>%
  mutate(
    month = 1 + (time - 1) %% 12,
    year = start_year + (time - 1) %/% 12,
    health_facility = factor(health_facility)
  )

ggplot(
  aes(
    x = month,
    y = cases,
    colour = health_facility,
    group = health_facility
  ),
  data = cases) +
  facet_wrap(~year) +
  geom_line(
    linewidth = 0.3
  ) + 
  theme_minimal() +
  scale_colour_discrete(
    guide = "none"
  ) +
  scale_x_continuous(breaks = 1:12)
```

and the prevalence survey results
```{r vis_data_3}
prev_surveys <- data$epi_data$prevalence_surveys %>%
  mutate(
    prevalence = n_positive / n_sampled
  )

ggplot() +
  geom_spatraster(data = pop_kenya_lores) +
  scale_fill_gradient(
    low = grey(0.9),
    high = grey(0.9),
    na.value = "transparent",
    guide = "none") +
  geom_point(
    aes(
      y = y,
      x = x,
      colour = prevalence
    ),
    data = prev_surveys,
    size = 3,
    shape = 16
  ) +
  scale_colour_gradient(
    labels = scales::label_percent(),
    low = "skyblue",
    high = "darkred"
  ) +
  xlab("") +
  ylab("") +
  theme_minimal()

```
