---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# epiwave.mapping

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Prototype R package and example code for combined semi-mechanistic spatio-temporal mapping of disease infection incidence simultaneously from spatially-aggregated (ie. at health facility level) clinical incidence timeseries, and targeted infection prevalence surveys at specific point locations. This is based on ideas in the epiwave R package and uses some functions implemented there, but largely uses different code that is more computationally convenient and practical for disease mapping applications.

## Installation

You can install the package from github, using the `remotes` package:

```{r install}
remotes::install_github("idem-lab/epiwave.mapping",
                        dependencies = TRUE)
```
